### 2015-11-25 DP3/analysis/Report.tex
finished DP3

\documentclass[english]{article}

\title{\vspace{-5.5ex} \bf Design Project 3: Dynamic Scheduling}
\author{{\bf Connor Smith} \\
	{\small Department of Electrical and Computer Engineering} \\
	{\small Brigham Young University, Provo, UT 84602} \\
	{\small connor.smith.256@gmail.com} \\
}

\usepackage{babel}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{float}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{setspace}

\date{\today}

\begin{document}
\doublespacing
\maketitle

\section{Introduction}
For decades, processors have used pipelining to exploit instruction-level parallelism. However, dependences among instructions as well as memory latency result in pipeline stalls and reduce throughput. Out-of-order processors seek to minimize the number of idle cycles by reordering the execution of instructions, a process known as dynamic scheduling. As long as this reordering is guaranteed to produce the same results as in-order execution, dynamic scheduling can minimize or hide stalls, increasing overall performance.

Many architectural parameters contribute to the organization of a processor. For a typical processor, these include the cache configuration, branch predictor, and number of functional units. For out-of-order processors, these also include the instruction queue size, the decode bandwidth, the RUU issue width and size, instruction commit bandwidth, and the load/store queue size. Each of these (and others) must be carefully chosen to maximize the desired characteristics. Since the design space is so large, it is necessary to use a methodology that is capable of considerably narrowing the space to determine the best configuration for a chosen market.

\section{Design Methodology}

\subsection{Base Configuration}
The design choices in this project reflect the target market of mid- to high-performance desktop computing. As a result, the primary characteristic of concern is overall performance. Since the cycle time is set to 600 MHz and does not vary from one design to the next, instructions per cycle (IPC) is a fair metric to use when comparing the performance of different configurations. Constraints were also placed on the average and maximum power consumption of the processor in an attempt to limit designs to those applicable to a typical desktop computer that does not require any special cooling.

The starting configuration is mostly based off of the default values in sim-outorder, with some modifications. The L1 and L2 cache configurations are based off those used in Intel's Haswell architecture and used in Design Project 1, and these remain constant across all designs explored in this project, since the focus of this project is on dynamic scheduling. The memory latency was fixed at 100ns, which corresponds to 60 cycles for a processory running at 600 MHz. Also, a perfect branch predictor was chosen for all of the designs. It should be noted that this limits the accuracy of the results; the reported IPC will be higher than in reality, and the reported power will likely be slightly lower. However, this choice eliminates a source of complexity, and the ranking of which designs are better should remain roughly the same. In addition, it is unlikely that the architecture would be utilized to its full potential when using an imperfect predictor, so the use of a perfect branch predictor should illustrate the full difference in performance between designs. Thus the end goal is less focused on presenting completely accurate metrics for a given design and more on determining which designs are better relative to  others. The parameters related to out-of-order execution in the starting configuration were set to be the defaults of sim-outorder. Table 1 shows a summary of the base configuration.

\begin{center}
    \singlespacing
	\captionof{table}{Baseline configuration}
	\vspace{.30cm}
	\begin{tabular}{ | l | l | l | l | l | l |}
	\hline
      L1 Capacity & 32KB \\ \hline
      L1 Associativity & 8 \\ \hline
      L2 Capacity & 128KB \\ \hline
      L2 Associativity & 8 \\ \hline
      Block Size & 64 \\ \hline
      Memory latency & 60 \\ \hline
      Branch prediction & perfect \\ \hline
	  IFQ size & 4 \\ \hline
      Decode width & 4 \\ \hline
      Issue width & 4 \\ \hline
      Commit width & 4 \\ \hline
      RUU size & 16 \\ \hline
      LSQ size & 8 \\ \hline
      I ALU & 4 \\ \hline
      I MULT & 1 \\ \hline
      FP ALU & 4 \\ \hline
      FP MULT & 1 \\ \hline
	\end{tabular}
\end{center}

\subsection{Parameter Variation}
Efficient parameter variation is necessary to narrow the design space. For each configuration that is the best so far, a single next parameter is chosen and varied. If the new configuration is better overall, then the parameter is further varied in the same direction until it is no longer beneficial. At this point, a new parameter is chosen and the process repeats until no further optimizations can be made. It is important to note that parameter values for a local maxima are not necessarily the same as for global maxima. In other words, the "best" value for a given parameter may change after altering a different parameter. Because of this, parameters were revisited at the end of the analysis to ensure that the chosen design really is the best so far.

The most important criterion for whether a given configuration change is beneficial is raw performance, measured in IPC. For many changes, however, the IPC increases slightly but at the cost of significantly more power dissipation. This tradeoff is evaluated and a determination is made whether it is worth it. Since the target market for these designs is desktop computing, a soft power limit was set at 150W. Both average and maximum power per cycle are used in these considerations. Average power is important for overall energy consumption, while maximum power is an important consideration due to heat dissipation issues. SimpleScalar offers multiple different power models, and the non-ideal conditional clocking (CC3) model was used for this project.

There are a few more metrics that were used to determine which parameter to change next. These are the fractions of time that the instruction queue, register update unit, and load/store queues were full. While these statistics are not directly useful for determining which configuration is best, they do provide a guide for whether greater performance would result from changing the parameter that corresponds to the full rate for a given unit.

\section{Configurations and Results}
\subsection{Preliminary Configurations}
The results for the baseline configuration are shown in Figures 1 and 2. As can be seen, the IPC varies wildly from one benchmark to the next, ranging from 0.162 to 2.72, with a mean of 1.330. The SPEC benchmarks are quite diverse, so in order for a configuration to be better, it must yield an increase in the average IPC (and hopefully across all benchmarks).

\begin{figure}[H]
	\centering
	\includegraphics[height=8cm]{baseline_ipc.png}
	\caption{Instructions per cycle by benchmark}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[height=8cm]{baseline_p.png}
	\caption{Average and maximum power by benchmark}
\end{figure}

Table 2 shows the average results for several different preliminary configurations, including the baseline. These configurations were analyzed to understand how much of an effect some of the parameters have on IPC and power. The "Double" configuration is the result of doubling each of the values, which yielded significant performance increases and kept power reasonable. "Max FUs" is the result of increasing the number of all the functional units to 8, which is the maximum value supported by SimpleScalar. Interestingly, this had a negligible effect on IPC but resulted in a spike in power, which illustrates that there are little to no structural hazards with the functional units in the baseline configuration. The "Min FUs" configuration corresponds to having only a single functional unit of each type. This caused a dramatic drop in performance, and was used to compute a soft minimum for IPC.

\begin{center}
	\captionof{table}{Preliminary Configuration Results}
	\vspace{.30cm}
	\begin{tabular}{ | l | l | l | l | l | l |}
		\hline
	  & Baseline & Double & Max FUs & Min FUs \\ \hline
      IPC & 1.330 & 1.896 & 1.367 & 0.788 \\ \hline
      Average Power & 18.078 & 31.540 & 22.185 & 12.290 \\ \hline
      Peak Power & 36.070 & 75.266 & 43.278 & 25.416 \\ \hline
	\end{tabular}
\end{center}

\subsection{Register Update Unit}
The first parameter that was changed was the register update unit (RUU), which is a combination of reservations stations and a reorder buffer (ROB). Based off the results of the baseline configuration, the RUU was full 50\% of the time. Increasing the bandwidth of the other units will not help much if the RUU is full, so the size was increased to 32 and 64. This decreased the full rate to roughly 1\%. The results for these configurations are shown in Table 3. IPC sees a modest improvement of 1.05x, while average and peak power increase by roughly 25\% each. There are clearly diminishing returns when increasing IPC.

\begin{center}
	\captionof{table}{Increasing RUU Size}
	\vspace{.30cm}
	\begin{tabular}{ | l | l | l | l | l | l |}
		\hline
	 RUU Size & 16 & 32 & 64 \\ \hline
     IPC & 1.330 & 1.386 & 1.389 \\ \hline
     Average Power & 18.078 & 20.384 & 22.561 \\ \hline
     Peak Power & 36.070 & 40.135 & 44.555 \\ \hline
	\end{tabular}
\end{center}

\subsection{Out-of-Order Bandwidth}
With a larger RUU, a good next choice of parameter to alter was the bandwidths of the out-of-order units: the fetch queue size, decode width, issue width, and commit width. These values were increased from 4 to 8. While they could be increased separately, it did not seem to make much sense to have a fetch queue size or commit width that were different, for example, so for the sake of simplicity each value was altered as a group. The results are shown in Table 4. As with increasing the RUU size, changing these parameters yielded a modest improvement of 1.05x to IPC. This time, however, average and peak power increased by 43\% and 78\%, respectively. These numbers are still well within the typical range for a desktop CPU, so the increase in IPC is worth it.

\begin{center}
	\captionof{table}{Increasing Out-of-order Unit Size/Bandwidth}
	\vspace{.30cm}
	\begin{tabular}{ | l | l | l | l | l | l |}
		\hline
	 Size/BW & 4 & 8 \\ \hline
     IPC & 1.389 & 1.463 \\ \hline
     Average Power & 22.561 & 26.045 \\ \hline
     Peak Power & 44.555 & 60.148 \\ \hline
	\end{tabular}
\end{center}

\subsection{Functional Units}
Using the architecture at this point, the processor can reorder and feed multiple instructions per cycle to the functional units, but there are still only a few FUs available. To reduce this congestion, the count of each functional unit was doubled, bringing the number of ALUs to 8 and multiply units to 2 for both integer and floating-point. Table 5 shows the results for this change. Yet again, this change resulted in a 1.05x boost in IPC. Average power increased by another 40\%, and peak power increased by 46\%. At this point, the power ceiling for desktop computing isn't too far off from the peak value of 88W reached by this new configuration.

\begin{center}
	\captionof{table}{Adding Functional Units}
	\vspace{.30cm}
	\begin{tabular}{ | l | l | l | l | l | l |}
		\hline
	 Number of FUs (ALU, MULT) & (4, 1) & (8, 2) \\ \hline
     IPC & 1.463 & 1.547 \\ \hline
     Average Power & 26.045 & 36.479 \\ \hline
     Peak Power & 60.148 & 87.705 \\ \hline
	\end{tabular}
\end{center}

\subsection{Load/Store Queue}
Thus far, not much has been done about latencies due to memory constraints. Indeed, at this point, the L/S queue is full 80\% of the time. Increasing the size of the load/store queue can help hide these limitations. The size of the queue was increased from 8 to 16, and the results are shown in Table 6. This caused the largest increase in IPC so far (1.36x). The increases in average and peak power (22\% and 40\%, respectively) were also some of the smallest of the changes so far. Increasing the size of the L/S queue was clearly a big win.

\begin{center}
	\captionof{table}{Adding Functional Units}
	\vspace{.30cm}
	\begin{tabular}{ | l | l | l | l | l | l |}
		\hline
	 L/S Queue Size & 8 & 16 & 32 & 64 \\ \hline
     IPC & 1.547 & 1.886 & 2.073 & 2.100\\ \hline
     Average Power & 36.479 & 40.853 & 43.485 & 44.376 \\ \hline
     Peak Power & 87.705 & 106.141 & 116.737 & 122.404 \\ \hline
	\end{tabular}
\end{center}

\subsection{Returning to Past Parameters}
At this point, there has been an overall IPC improvement of 1.58x, at the cost of a 2.45x increase in average power and 3.4x gain in peak power. With a peak power of over 120W, any further changes must have dramatic performance increases to justify increasing the power consumption any further. Since the number of ALUs was already at the maximum value that SimpleScalar supports, only the out-of-order width parameters and the RUU size were left to revisit. Table 7 shows the results of these changes. Increasing the RUU size does yield a performance increase, but at the cost of increasing the peak power consumption to a whopping 206W, which simply cannot be justified. In addition, increasing the issue/dispatch/commit widths again actually resulted in a net decrease in IPC while still increasing power consumption drastically.

\begin{center}
	\captionof{table}{Revisiting Past Parameters}
	\vspace{.30cm}
	\begin{tabular}{ | l | l | l | l | l | l |}
		\hline
	 & Best So Far & Increasing RUU Size & Increasing width  \\ \hline
     IPC & 2.100 & 2.373 & 2.065 \\ \hline
     Average Power & 44.376 & 61.814 & 63.878 \\ \hline
     Peak Power & 122.404& 206.089 & 185.695  \\ \hline
	\end{tabular}
\end{center}

\section{Conclusion}
The final configuration is shown in Table 8, and a comparison of IPC for each benchmark is shown in Figure 3. There is a clear and significant improvement in performance; no benchmark performs worse under the new configuration, and while some see only marginal benefit, those with high starting IPC see a dramatic improvement.

\begin{center}
    \singlespacing
	\captionof{table}{Baseline configuration}
	\vspace{.30cm}
	\begin{tabular}{ | l | l | l | l | l | l |}
	\hline
	  IFQ size & 8 \\ \hline
      Decode width & 8 \\ \hline
      Issue width & 8 \\ \hline
      Commit width & 8 \\ \hline
      RUU size & 64 \\ \hline
      LSQ size & 64 \\ \hline
      I ALU & 8 \\ \hline
      I MULT & 2 \\ \hline
      FP ALU & 8 \\ \hline
      FP MULT & 2 \\ \hline
	\end{tabular}
\end{center}

\begin{figure}[H]
	\centering
	\includegraphics[height=8cm]{final.png}
	\caption{Baseline vs. Final configurations}
\end{figure}

Designing an out-of-order processor is clearly much more involved than designing a simple in-order one. Each of the different parameters can affect each other significantly, and even after narrowing the design space considerably, there are still a huge amount of configurations available. Only a few such designs could be tested within the scope of this project, but they should be representative. As with most designs in computer architecture, the configurations in this project were full of trade-offs. If the target market were chosen differently, the best design would be markedly different from the one selected here. If a lower IPC were acceptable, the power consumption could be drastically lower. Likewise, if power consumption wasn't as much of a concern, more performance could have been squeezed out of the designs. OVerall, the final design strikes a good balance between performance and power consumption, which is excellent for the chosen market.

\end{document}

### 2015-11-25 DP3/scripts/csv.py

@@ -1,0 +1,58 @@
+#!/u/cjs1993/python3/bin/python3
+
+import sys, glob, os, re, statistics
+
+config_dir = '../configs'
+results_dir = '../results'
+blacklist_file = '../analysis/blacklist'
+
+def reduce(config, metric, func):
+  values = []
+  pattern = re.compile(metric + ' ')
+  for benchmark in glob.iglob(results_dir + '/' + config + '/*.eio.gz.out'):
+    for line in open(benchmark):
+      if pattern.match(line):
+        #print(benchmark, line)
+        value = float(re.findall(r'\s+\d*\.?\d+', line)[0])
+        values.append(value)
+  #print(config, values)
+  return func(values)
+
+method = sys.argv[2]
+if method == "mean":
+  func = statistics.mean
+elif method == "median":
+  func = statistics.median
+elif method == "min":
+  func = min
+elif method == "max":
+  func = max
+
+configs = [os.path.splitext(os.path.basename(filename))[0] for filename in glob.iglob('../configs/*.values')]
+blacklist = [line.replace('\n', '') for line in open(blacklist_file)]
+configs = [config for config in configs if not config in blacklist]
+configs.sort()
+
+columns = '_,'
+
+# column titles
+for config in configs:
+  columns += config + ','
+columns = columns[:-1]
+
+print(columns)
+
+# for each row/metric
+for metric in open(sys.argv[1]):
+  metric = metric.replace('\n', '')
+  row = metric + ','
+
+  for config in configs:
+    value = reduce(config, metric, func)
+    row += format(value, '.5f') + ','
+  
+  row = row[:-1]
+  print(row)
+
+print('\n')
+

### 2015-11-25 DP3/analysis/Report.tex
fixed some grammatical errors

@@
 These values were increased from 4 to 8.
-While they could be increased separately, it did not seem to make much sense to have a fetch queue size or commit width that were different, for example, so for the sake of simplicity each value was altered as a group.
+While they could be increased separately, it did not seem to make much sense to do so.
+Thus for the sake of simplicity each value was altered as a group.
 The results are shown in Table 4.
@@
 Likewise, if power consumption wasn't as much of a concern, more performance could have been squeezed out of the designs.
-OVerall, the final design strikes a good balance between performance and power consumption, which is excellent for the chosen market.
+Overall, the final design strikes a good balance between performance and power consumption, which is excellent for the chosen market.
 

### 2015-12-05 FinalProject/Tex/acmsmall/sample1.tex
added template for acmsmall

% v2-acmsmall-sample.tex, dated March 6 2012
% This is a sample file for ACM small trim journals
%
% Compilation using 'acmsmall.cls' - version 1.3 (March 2012), Aptara Inc.
% (c) 2010 Association for Computing Machinery (ACM)
%
% Questions/Suggestions/Feedback should be addressed to => "acmtexsupport@aptaracorp.com".
% Users can also go through the FAQs available on the journal's submission webpage.
%
% Steps to compile: latex, bibtex, latex latex
%
% For tracking purposes => this is v1.3 - March 2012

\documentclass[prodmode,acmtecs]{acmsmall} % Aptara syntax

% Package to generate and customize Algorithm as per ACM style
\usepackage[ruled]{algorithm2e}
\renewcommand{\algorithmcfname}{ALGORITHM}
\SetAlFnt{\small}
\SetAlCapFnt{\small}
\SetAlCapNameFnt{\small}
\SetAlCapHSkip{0pt}
\IncMargin{-\parindent}

% Metadata Information
\acmVolume{9}
\acmNumber{4}
\acmArticle{39}
\acmYear{2010}
\acmMonth{3}

% Copyright
\setcopyright{rightsretained}

% DOI
\doi{0000001.0000001}

%ISSN
\issn{1234-56789}

% Document starts
\begin{document}

% Page heads
\markboth{G. Zhou et al.}{A Multifrequency MAC Specially Designed for WSN Applications}

% Title portion
\title{A Multifrequency MAC Specially Designed for Wireless Sensor Network Applications}
\author{GANG ZHOU
\affil{College of William and Mary}
YAFENG WU
\affil{University of Virginia}
TING YAN
\affil{Eaton Innovation Center}
TIAN HE
\affil{University of Minnesota}
CHENGDU HUANG
\affil{Google}
JOHN A. STANKOVIC
\affil{University of Virginia}
TAREK F. ABDELZAHER
\affil{University of Illinois at Urbana-Champaign}}
% NOTE! Affiliations placed here should be for the institution where the
%       BULK of the research was done. If the author has gone to a new
%       institution, before publication, the (above) affiliation should NOT be changed.
%       The authors 'current' address may be given in the "Author's addresses:" block (below).
%       So for example, Mr. Abdelzaher, the bulk of the research was done at UIUC, and he is
%       currently affiliated with NASA.

\begin{abstract}
Multifrequency media access control has been well understood in
general wireless ad hoc networks, while in wireless sensor networks,
researchers still focus on single frequency solutions. In wireless
sensor networks, each device is typically equipped with a single
radio transceiver and applications adopt much smaller packet sizes
compared to those in general wireless ad hoc networks. Hence, the
multifrequency MAC protocols proposed for general wireless ad hoc
networks are not suitable for wireless sensor network applications,
which we further demonstrate through our simulation experiments. In
this article, we propose MMSN, which takes advantage of
multifrequency availability while, at the same time, takes into
consideration the restrictions of wireless sensor networks. Through
extensive experiments, MMSN exhibits the prominent ability to utilize
parallel transmissions among neighboring nodes. When multiple physical
frequencies are available, it also achieves increased energy
efficiency, demonstrating the ability to work against radio
interference and the tolerance to a wide range of measured time
synchronization errors.
\end{abstract}

%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below.
%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10010520.10010553.10010562</concept_id>
  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010575.10010755</concept_id>
  <concept_desc>Computer systems organization~Redundancy</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010553.10010554</concept_id>
  <concept_desc>Computer systems organization~Robotics</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>10003033.10003083.10003095</concept_id>
  <concept_desc>Networks~Network reliability</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Computer systems organization~Embedded systems}
\ccsdesc[300]{Computer systems organization~Redundancy}
\ccsdesc{Computer systems organization~Robotics}
\ccsdesc[100]{Networks~Network reliability}

%
% End generated code
%

\terms{Design, Algorithms, Performance}

\keywords{Wireless sensor networks, media access control,
multi-channel, radio interference, time synchronization}

\acmformat{Gang Zhou, Yafeng Wu, Ting Yan, Tian He, Chengdu Huang, John A. Stankovic,
and Tarek F. Abdelzaher, 2010. A multifrequency MAC specially
designed for  wireless sensor network applications.}
% At a minimum you need to supply the author names, year and a title.
% IMPORTANT:
% Full first names whenever they are known, surname last, followed by a period.
% In the case of two authors, 'and' is placed between them.
% In the case of three or more authors, the serial comma is used, that is, all author names
% except the last one but including the penultimate author's name are followed by a comma,
% and then 'and' is placed before the final author's name.
% If only first and middle initials are known, then each initial
% is followed by a period and they are separated by a space.
% The remaining information (journal title, volume, article number, date, etc.) is 'auto-generated'.

\begin{bottomstuff}
This work is supported by the National Science Foundation, under
grant CNS-0435060, grant CCR-0325197 and grant EN-CS-0329609.

Author's addresses: G. Zhou, Computer Science Department,
College of William and Mary; Y. Wu  {and} J. A. Stankovic,
Computer Science Department, University of Virginia; T. Yan,
Eaton Innovation Center; T. He, Computer Science Department,
University of Minnesota; C. Huang, Google; T. F. Abdelzaher,
(Current address) NASA Ames Research Center, Moffett Field, California 94035.
\end{bottomstuff}

\maketitle

\section{Introduction}

As a new technology, Wireless Sensor Networks (WSNs) has a wide
range of applications [Culler 2001,Bahl 2002,Akyildiz 2001], including
environment monitoring, smart buildings, medical care, industrial and
military applications. Among them, a recent trend is to develop
commercial sensor networks that require pervasive sensing of both
environment and human beings, for example, assisted living
[Akyildiz 2002,Harvard 2001,CROSSBOW] and smart homes
[Harvard 2001,Adya 2001,CROSSBOW].
% quote
\begin{quote}
``For these applications, sensor devices are incorporated into human
cloths [Natarajan 2001,Zhou 2006,Bahl 2002,Adya 2001] for monitoring
health related information like EKG readings, fall detection, and voice recognition".
\end{quote}
While collecting all these multimedia information
[Akyildiz 2002] requires a high network throughput, off-the-shelf
sensor devices only provide very limited bandwidth in a single
channel: 19.2Kbps in MICA2 [Bahl 2002] and 250Kbps in MICAz.

In this article, we propose MMSN, abbreviation for Multifrequency
Media access control for wireless Sensor Networks. The main
contributions of this work can be summarized as follows.
% itemize
\begin{itemize}
\item To the best of our knowledge, the MMSN protocol is the first
multifrequency MAC protocol especially designed for WSNs, in which
each device is equipped with a single radio transceiver and
the MAC layer packet size is very small.
\item Instead of using pairwise RTS/CTS frequency negotiation
[Adya 2001,Culler 2001; Tzamaloukas 2001; Zhou 2006],
we propose lightweight frequency assignments, which are good choices
for many deployed comparatively static WSNs.
\item We develop new toggle transmission and snooping techniques to
enable a single radio transceiver in a sensor device to achieve
scalable performance, avoiding the nonscalable ``one
control channel + multiple data channels'' design [Natarajan 2001].
\end{itemize}

% Head 1
\section{MMSN Protocol}

% Head 2
\subsection{Frequency Assignment}

We propose a suboptimal distribution to be used by each node, which is
easy to compute and does not depend on the number of competing
nodes. A natural candidate is an increasing geometric sequence, in
which
% Numbered Equation
\begin{equation}
\label{eqn:01}
P(t)=\frac{b^{\frac{t+1}{T+1}}-b^{\frac{t}{T+1}}}{b-1},
\end{equation}
where $t=0,{\ldots}\,,T$, and $b$ is a number greater than $1$.

In our algorithm, we use the suboptimal approach for simplicity and
generality. We need to make the distribution of the selected back-off
time slice at each node conform to what is shown in Equation
(\ref{eqn:01}). It is implemented as follows: First, a random
variable $\alpha$ with a uniform distribution within the interval
$(0, 1)$ is generated on each node, then time slice $i$ is selected
according to the following equation:
% Unnumbered Equation
\[
i=\lfloor(T+1)\log_b[\alpha(b-1)+1]\rfloor.
\]
It can be easily proven that the distribution of $i$ conforms to Equation
(\ref{eqn:01}).

So protocols [Bahl 2002,Culler 2001,Zhou 2006,Adya 2001,Culler 2001;
Tzamaloukas-01; Akyildiz-01] that use RTS/CTS
controls\footnote{RTS/CTS controls are required to be implemented by
802.11-compliant devices. They can be used as an optional mechanism
to avoid Hidden Terminal Problems in the 802.11 standard and
protocols based on those similar to [Akyildiz 2001] and
[Adya 2001].} for frequency negotiation and reservation are not
suitable for WSN applications, even though they exhibit good
performance in general wireless ad hoc
networks.

% Head 3
\subsubsection{Exclusive Frequency Assignment}

In exclusive frequency assignment, nodes first exchange their IDs
among two communication hops so that each node knows its two-hop
neighbors' IDs. In the second broadcast, each node beacons all
neighbors' IDs it has collected during the first broadcast period.

% Head 4
\paragraph{Eavesdropping}

Even though the even selection scheme leads to even sharing of
available frequencies among any two-hop neighborhood, it involves a
number of two-hop broadcasts. To reduce the communication cost, we
propose a lightweight eavesdropping scheme.

\subsection{Basic Notations}

As Algorithm~\ref{alg:one} states, for each frequency
number, each node calculates a random number (${\textit{Rnd}}_{\alpha}$) for
itself and a random number (${\textit{Rnd}}_{\beta}$) for each of its two-hop
neighbors with the same pseudorandom number generator.
% Algorithm
\begin{algorithm}[t]
\SetAlgoNoLine
\KwIn{Node $\alpha$'s ID ($ID_{\alpha}$), and node $\alpha$'s
neighbors' IDs within two communication hops.}
\KwOut{The frequency number ($FreNum_{\alpha}$) node $\alpha$ gets assigned.}
$index$ = 0; $FreNum_{\alpha}$ = -1\;
\Repeat{$FreNum_{\alpha} > -1$}{
        $Rnd_{\alpha}$ = Random($ID_{\alpha}$, $index$)\;
        $Found$ = $TRUE$\;
        \For{each node $\beta$ in $\alpha$'s two communication hops
    }{
      $Rnd_{\beta}$ = Random($ID_{\beta}$, $index$)\;
      \If{($Rnd_{\alpha} < Rnd_{\beta}$) \text{or} ($Rnd_{\alpha}$ ==
          $Rnd_{\beta}$ \text{and} $ID_{\alpha} < ID_{\beta}$)\;
      }{
        $Found$ = $FALSE$; break\;
      }
        }
     \eIf{$Found$}{
           $FreNum_{\alpha}$ = $index$\;
         }{
           $index$ ++\;
     }
      }
\caption{Frequency Number Computation}
\label{alg:one}
\end{algorithm}

Bus masters are divided into two disjoint sets, $\mathcal{M}_{RT}$
and $\mathcal{M}_{NRT}$.
% description
\begin{description}
\item[RT Masters]
$\mathcal{M}_{RT}=\{ \vec{m}_{1},\dots,\vec{m}_{n}\}$ denotes the
$n$ RT masters issuing real-time constrained requests. To model the
current request issued by an $\vec{m}_{i}$ in $\mathcal{M}_{RT}$,
three parameters---the recurrence time $(r_i)$, the service cycle
$(c_i)$, and the relative deadline $(d_i)$---are used, with their
relationships.
\item[NRT Masters]
$\mathcal{M}_{NRT}=\{ \vec{m}_{n+1},\dots,\vec{m}_{n+m}\}$ is a set
of $m$ masters issuing nonreal-time constrained requests. In our
model, each $\vec{m}_{j}$ in $\mathcal{M}_{NRT}$ needs only one
parameter, the service cycle, to model the current request it
issues.
\end{description}

Here, a question may arise, since each node has a global ID. Why
don't we just map nodes' IDs within two hops into a group of
frequency numbers and assign those numbers to all nodes within two
hops?

\section{Simulator}
\label{sec:sim}

If the model checker requests successors of a state which are not
created yet, the state space uses the simulator to create the
successors on-the-fly. To create successor states the simulator
conducts the following steps.
% enumerate
\begin{enumerate}
\item Load state into microcontroller model.
\item Determine assignments needed for resolving nondeterminism.
\item For each assignment.
      \begin{enumerate}
      \item either call interrupt handler or simulate effect of next instruction, or
      \item evaluate truth values of atomic propositions.
      \end{enumerate}
\item Return resulting states.
\end{enumerate}
Figure~\ref{fig:one} shows a typical microcontroller C program that
controls an automotive power window lift. The program is one of the
programs used in the case study described in Section~\ref{sec:sim}.
At first sight, the programs looks like an ANSI~C program. It
contains function calls, assignments, if clauses, and while loops.
% Figure
\begin{figure}
\centerline{\includegraphics{acmsmall-mouse}}
\caption{Code before preprocessing.}
\label{fig:one}
\end{figure}

\subsection{Problem Formulation}

The objective of variable coalescence-based offset assignment is to find
both the coalescence scheme and the MWPC on the coalesced graph. We start
with a few definitions and lemmas for variable coalescence.

% Enunciations
\begin{definition}[Coalesced Node (C-Node)]A C-node is a set of
live ranges (webs) in the AG or IG that are coalesced. Nodes within the same
C-node cannot interfere with each other on the IG. Before any coalescing is
done, each live range is a C-node by itself.
\end{definition}

\begin{definition}[C-AG (Coalesced Access Graph)]The C-AG is the access
graph after node coalescence, which is composed of all C-nodes and C-edges.
\end{definition}

\begin{lemma}
The C-MWPC problem is NP-complete.
\end{lemma}
\begin{proof} C-MWPC can be easily reduced to the MWPC problem assuming a
coalescence graph without any edge or a fully connected interference graph.
Therefore, each C-node is an uncoalesced live range after value separation
and C-PC is equivalent to PC. A fully connected interference graph is made
possible when all live ranges interfere with each other. Thus, the C-MWPC
problem is NP-complete.
\end{proof}

\begin{lemma}[Lemma Subhead]The solution to the C-MWPC problem is no
worse than the solution to the MWPC.
\end{lemma}
\begin{proof}
Simply, any solution to the MWPC is also a solution to the
C-MWPC. But some solutions to C-MWPC may not apply to the MWPC (if any
coalescing were made).
\end{proof}

\section{Performance Evaluation}

During all the experiments, the Geographic Forwarding (GF)
[Akyildiz 2001] routing protocol is used. GF exploits geographic
information of nodes and conducts local data-forwarding to achieve
end-to-end routing. Our simulation is
configured according to the settings in
Table~\ref{tab:one}. Each run lasts for 2 minutes and
repeated 100 times. For each data value we present in the results,
we also give its 90\% confidence interval.
% Table
\begin{table}%
\tbl{Simulation Configuration\label{tab:one}}{%
\begin{tabular}{|l|l|}
\hline
TERRAIN{$^a$}   & (200m$\times$200m) Square\\\hline
Node Number     & 289\\\hline
Node Placement  & Uniform\\\hline
Application     & Many-to-Many/Gossip CBR Streams\\\hline
Payload Size    & 32 bytes\\\hline
Routing Layer   & GF\\\hline
MAC Layer       & CSMA/MMSN\\\hline
Radio Layer     & RADIO-ACCNOISE\\\hline
Radio Bandwidth & 250Kbps\\\hline
Radio Range     & 20m--45m\\\hline
\end{tabular}}
\begin{tabnote}%
\Note{Source:}{This is a table
sourcenote. This is a table sourcenote. This is a table
sourcenote.}
\vskip2pt
\Note{Note:}{This is a table footnote.}
\tabnoteentry{$^a$}{This is a table footnote. This is a
table footnote. This is a table footnote.}
\end{tabnote}%
\end{table}%

\section{Conclusions}

In this article, we develop the first multifrequency MAC protocol for
WSN applications in which each device adopts a
single radio transceiver. The different MAC design requirements for
WSNs and general wireless ad-hoc networks are
compared, and a complete WSN multifrequency MAC design (MMSN) is
put forth. During the MMSN design, we analyze and evaluate different
choices for frequency assignments and also discuss the nonuniform
back-off algorithms for the slotted media access design.

% Start of "Sample References" section

\section{Typical references in new ACM Reference Format}
A paginated journal article \cite{Abril07}, an enumerated
journal article \cite{Cohen07}, a reference to an entire issue \cite{JCohen96},
a monograph (whole book) \cite{Kosiur01}, a monograph/whole book in a series (see 2a in spec. document)
\cite{Harel79}, a divisible-book such as an anthology or compilation \cite{Editor00}
followed by the same example, however we only output the series if the volume number is given
\cite{Editor00a} (so Editor00a's series should NOT be present since it has no vol. no.),
a chapter in a divisible book \cite{Spector90}, a chapter in a divisible book
in a series \cite{Douglass98}, a multi-volume work as book \cite{Knuth97},
an article in a proceedings (of a conference, symposium, workshop for example)
(paginated proceedings article) \cite{Andler79}, a proceedings article
with all possible elements \cite{Smith10}, an example of an enumerated
proceedings article \cite{VanGundy07},
an informally published work \cite{Harel78}, a doctoral dissertation \cite{Clarkson85},
a master's thesis: \cite{anisi03}, an online document / world wide web resource \cite{Thornburg01}, \cite{Ablamowicz07},
\cite{Poker06}, a video game (Case 1) \cite{Obama08} and (Case 2) \cite{Novak03}
and \cite{Lee05} and (Case 3) a patent \cite{JoeScientist001},
work accepted for publication \cite{rous08}, 'YYYYb'-test for prolific author
\cite{SaeediMEJ10} and \cite{SaeediJETC10}. Other cites might contain
'duplicate' DOI and URLs (some SIAM articles) \cite{Kirschmer:2010:AEI:1958016.1958018}.
Boris / Barbara Beeton: multi-volume works as books
\cite{MR781536} and \cite{MR781537}.

% Appendix
\appendix
\section*{APPENDIX}
\setcounter{section}{1}
In this appendix, we measure
the channel switching time of Micaz [CROSSBOW] sensor devices.
In our experiments, one mote alternatingly switches between Channels
11 and 12. Every time after the node switches to a channel, it sends
out a packet immediately and then changes to a new channel as soon
as the transmission is finished. We measure the
number of packets the test mote can send in 10 seconds, denoted as
$N_{1}$. In contrast, we also measure the same value of the test
mote without switching channels, denoted as $N_{2}$. We calculate
the channel-switching time $s$ as
\begin{eqnarray}%
s=\frac{10}{N_{1}}-\frac{10}{N_{2}}. \nonumber
\end{eqnarray}%
By repeating the experiments 100 times, we get the average
channel-switching time of Micaz motes: 24.3$\mu$s.

\appendixhead{ZHOU}

% Acknowledgments
\begin{acks}
The authors would like to thank Dr. Maura Turolla of Telecom
Italia for providing specifications about the application scenario.
\end{acks}

% Bibliography
\bibliographystyle{ACM-Reference-Format-Journals}
\bibliography{acmsmall-sample-bibfile}
                             % Sample .bib file with references that match those in
                             % the 'Specifications Document (V1.5)' as well containing
                             % 'legacy' bibs and bibs with 'alternate codings'.
                             % Gerry Murray - March 2012

% History dates
\received{February 2007}{March 2009}{June 2009}

% Electronic Appendix
\elecappendix

\medskip

\section{This is an example of Appendix section head}

Channel-switching time is measured as the time length it takes for
motes to successfully switch from one channel to another. This
parameter impacts the maximum network throughput, because motes
cannot receive or send any packet during this period of time, and it
also affects the efficiency of toggle snooping in MMSN, where motes
need to sense through channels rapidly.

By repeating experiments 100 times, we get the average
channel-switching time of Micaz motes: 24.3 $\mu$s. We then conduct
the same experiments with different Micaz motes, as well as
experiments with the transmitter switching from Channel 11 to other
channels. In both scenarios, the channel-switching time does not have
obvious changes. (In our experiments, all values are in the range of
23.6 $\mu$s to 24.9 $\mu$s.)

\section{Appendix section head}

The primary consumer of energy in WSNs is idle listening. The key to
reduce idle listening is executing low duty-cycle on nodes. Two
primary approaches are considered in controlling duty-cycles in the
MAC layer.

\end{document}
% End of v2-acmsmall-sample.tex (March 2012) - Gerry Murray, ACM

### 2015-12-05 FinalProject/Tex/paper/paper.tex
set up final paper directory, cleaned up errors

--- 2015-12-05 FinalProject/Tex/acmsmall/sample1.tex
@@
-% v2-acmsmall-sample.tex, dated March 6 2012
-% This is a sample file for ACM small trim journals
-%
-% Compilation using 'acmsmall.cls' - version 1.3 (March 2012), Aptara Inc.
-% (c) 2010 Association for Computing Machinery (ACM)
-%
-% Questions/Suggestions/Feedback should be addressed to => "acmtexsupport@aptaracorp.com".
-% Users can also go through the FAQs available on the journal's submission webpage.
-%
-% Steps to compile: latex, bibtex, latex latex
-%
-% For tracking purposes => this is v1.3 - March 2012
-
 \documentclass[prodmode,acmtecs]{acmsmall} % Aptara syntax
@@
 % Copyright
+%\setcopyright{acmcopyright}
+%\setcopyright{acmlicensed}
 \setcopyright{rightsretained}
+%\setcopyright{usgov}
+%\setcopyright{usgovmixed}
+%\setcopyright{cagov}
+%\setcopyright{cagovmixed}
 
@@
 
-\terms{Design, Algorithms, Performance}
+% We no longer use \terms command
+%\terms{Design, Algorithms, Performance}
 
@@
 \begin{figure}
-\centerline{\includegraphics{acmsmall-mouse}}
+% \centerline{\includegraphics{acmsmall-mouse}}
 \caption{Code before preprocessing.}
@@
 \section{Typical references in new ACM Reference Format}
-A paginated journal article \cite{Abril07}, an enumerated
-journal article \cite{Cohen07}, a reference to an entire issue \cite{JCohen96},
-a monograph (whole book) \cite{Kosiur01}, a monograph/whole book in a series (see 2a in spec. document)
-\cite{Harel79}, a divisible-book such as an anthology or compilation \cite{Editor00}
+A paginated journal article \cite{test1}, an enumerated
+journal article \cite{test1}, a reference to an entire issue \cite{test1},
+a monograph (whole book) \cite{test1}, a monograph/whole book in a series (see 2a in spec. document)
+\cite{test1}, a divisible-book such as an anthology or compilation \cite{test1}
 followed by the same example, however we only output the series if the volume number is given
-\cite{Editor00a} (so Editor00a's series should NOT be present since it has no vol. no.),
-a chapter in a divisible book \cite{Spector90}, a chapter in a divisible book
-in a series \cite{Douglass98}, a multi-volume work as book \cite{Knuth97},
+\cite{test1} (so Editor00a's series should NOT be present since it has no vol. no.),
+a chapter in a divisible book \cite{test1}, a chapter in a divisible book
+in a series \cite{test1}, a multi-volume work as book \cite{test1},
 an article in a proceedings (of a conference, symposium, workshop for example)
-(paginated proceedings article) \cite{Andler79}, a proceedings article
-with all possible elements \cite{Smith10}, an example of an enumerated
-proceedings article \cite{VanGundy07},
-an informally published work \cite{Harel78}, a doctoral dissertation \cite{Clarkson85},
-a master's thesis: \cite{anisi03}, an online document / world wide web resource \cite{Thornburg01}, \cite{Ablamowicz07},
-\cite{Poker06}, a video game (Case 1) \cite{Obama08} and (Case 2) \cite{Novak03}
-and \cite{Lee05} and (Case 3) a patent \cite{JoeScientist001},
-work accepted for publication \cite{rous08}, 'YYYYb'-test for prolific author
-\cite{SaeediMEJ10} and \cite{SaeediJETC10}. Other cites might contain
-'duplicate' DOI and URLs (some SIAM articles) \cite{Kirschmer:2010:AEI:1958016.1958018}.
+(paginated proceedings article) \cite{test1}, a proceedings article
+with all possible elements \cite{test1}, an example of an enumerated
+proceedings article \cite{test1},
+an informally published work \cite{test1}, a doctoral dissertation \cite{test1},
+a master's thesis: \cite{test1}, an online document / world wide web resource \cite{test1}, \cite{test1},
+\cite{test1}, a video game (Case 1) \cite{test1} and (Case 2) \cite{test1}
+and \cite{test1} and (Case 3) a patent \cite{test1},
+work accepted for publication \cite{test1}, 'YYYYb'-test for prolific author
+\cite{test1} and \cite{test1}. Other cites might contain
+'duplicate' DOI and URLs (some SIAM articles) \cite{test1}.
 Boris / Barbara Beeton: multi-volume works as books
-\cite{MR781536} and \cite{MR781537}.
+\cite{test1} and \cite{test1}.
 
@@
 \bibliographystyle{ACM-Reference-Format-Journals}
-\bibliography{acmsmall-sample-bibfile}
+\bibliography{paper-bibfile}
                              % Sample .bib file with references that match those in

### 2015-12-05 FinalProject/Tex/paper/paper.tex
set up paper formatting

@@
 
-% Metadata Information
-\acmVolume{9}
-\acmNumber{4}
-\acmArticle{39}
-\acmYear{2010}
-\acmMonth{3}
+% % Metadata Information
+\acmVolume{000}
+\acmNumber{000}
+\acmArticle{000}
+\acmYear{2015}
+\acmMonth{12}
 
@@
 % Page heads
-\markboth{G. Zhou et al.}{A Multifrequency MAC Specially Designed for WSN Applications}
+\markboth{C. Smith}{TODO A Survey Paper}
 
 % Title portion
-\title{A Multifrequency MAC Specially Designed for Wireless Sensor Network Applications}
-\author{GANG ZHOU
-\affil{College of William and Mary}
-YAFENG WU
-\affil{University of Virginia}
-TING YAN
-\affil{Eaton Innovation Center}
-TIAN HE
-\affil{University of Minnesota}
-CHENGDU HUANG
-\affil{Google}
-JOHN A. STANKOVIC
-\affil{University of Virginia}
-TAREK F. ABDELZAHER
-\affil{University of Illinois at Urbana-Champaign}}
-% NOTE! Affiliations placed here should be for the institution where the
-%       BULK of the research was done. If the author has gone to a new
-%       institution, before publication, the (above) affiliation should NOT be changed.
-%       The authors 'current' address may be given in the "Author's addresses:" block (below).
-%       So for example, Mr. Abdelzaher, the bulk of the research was done at UIUC, and he is
-%       currently affiliated with NASA.
+\title{TODO A Survey Paper}
+\author{Connor Smith
+\affil{Brigham Young University}}
 
 \begin{abstract}
-Multifrequency media access control has been well understood in
-general wireless ad hoc networks, while in wireless sensor networks,
-researchers still focus on single frequency solutions. In wireless
-sensor networks, each device is typically equipped with a single
-radio transceiver and applications adopt much smaller packet sizes
-compared to those in general wireless ad hoc networks. Hence, the
-multifrequency MAC protocols proposed for general wireless ad hoc
-networks are not suitable for wireless sensor network applications,
-which we further demonstrate through our simulation experiments. In
-this article, we propose MMSN, which takes advantage of
-multifrequency availability while, at the same time, takes into
-consideration the restrictions of wireless sensor networks. Through
-extensive experiments, MMSN exhibits the prominent ability to utilize
-parallel transmissions among neighboring nodes. When multiple physical
-frequencies are available, it also achieves increased energy
-efficiency, demonstrating the ability to work against radio
-interference and the tolerance to a wide range of measured time
-synchronization errors.
+This is my abstract.
 \end{abstract}
 
-%
-% The code below should be generated by the tool at
-% http://dl.acm.org/ccs.cfm
-% Please copy and paste the code instead of the example below.
-%
-\begin{CCSXML}
-<ccs2012>
- <concept>
-  <concept_id>10010520.10010553.10010562</concept_id>
-  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
-  <concept_significance>500</concept_significance>
- </concept>
- <concept>
-  <concept_id>10010520.10010575.10010755</concept_id>
-  <concept_desc>Computer systems organization~Redundancy</concept_desc>
-  <concept_significance>300</concept_significance>
- </concept>
- <concept>
-  <concept_id>10010520.10010553.10010554</concept_id>
-  <concept_desc>Computer systems organization~Robotics</concept_desc>
-  <concept_significance>100</concept_significance>
- </concept>
- <concept>
-  <concept_id>10003033.10003083.10003095</concept_id>
-  <concept_desc>Networks~Network reliability</concept_desc>
-  <concept_significance>100</concept_significance>
- </concept>
-</ccs2012>
-\end{CCSXML}
+\keywords{TODO, General purpose graphics processing unit, heterogeneous systems,
+warp size}
 
-\ccsdesc[500]{Computer systems organization~Embedded systems}
-\ccsdesc[300]{Computer systems organization~Redundancy}
-\ccsdesc{Computer systems organization~Robotics}
-\ccsdesc[100]{Networks~Network reliability}
-
-%
-% End generated code
-%
-
-% We no longer use \terms command
-%\terms{Design, Algorithms, Performance}
-
-\keywords{Wireless sensor networks, media access control,
-multi-channel, radio interference, time synchronization}
-
-\acmformat{Gang Zhou, Yafeng Wu, Ting Yan, Tian He, Chengdu Huang, John A. Stankovic,
-and Tarek F. Abdelzaher, 2010. A multifrequency MAC specially
-designed for  wireless sensor network applications.}
-% At a minimum you need to supply the author names, year and a title.
-% IMPORTANT:
-% Full first names whenever they are known, surname last, followed by a period.
-% In the case of two authors, 'and' is placed between them.
-% In the case of three or more authors, the serial comma is used, that is, all author names
-% except the last one but including the penultimate author's name are followed by a comma,
-% and then 'and' is placed before the final author's name.
-% If only first and middle initials are known, then each initial
-% is followed by a period and they are separated by a space.
-% The remaining information (journal title, volume, article number, date, etc.) is 'auto-generated'.
+\acmformat{Connor Smith, 2015. TODO A Survey Paper.}
 
 \begin{bottomstuff}
-This work is supported by the National Science Foundation, under
-grant CNS-0435060, grant CCR-0325197 and grant EN-CS-0329609.
+This work is supported by the National Imaginary Science Foundation, under
+grant CNS-0123456.
 
-Author's addresses: G. Zhou, Computer Science Department,
-College of William and Mary; Y. Wu  {and} J. A. Stankovic,
-Computer Science Department, University of Virginia; T. Yan,
-Eaton Innovation Center; T. He, Computer Science Department,
-University of Minnesota; C. Huang, Google; T. F. Abdelzaher,
-(Current address) NASA Ames Research Center, Moffett Field, California 94035.
+Author's addresses: C. Smith, Electrical and Computer Engineering Department,
+Brigham Young University.
 \end{bottomstuff}
@@
 
-\section{Conclusions}
-
-In this article, we develop the first multifrequency MAC protocol for
-WSN applications in which each device adopts a
-single radio transceiver. The different MAC design requirements for
-WSNs and general wireless ad-hoc networks are
-compared, and a complete WSN multifrequency MAC design (MMSN) is
-put forth. During the MMSN design, we analyze and evaluate different
-choices for frequency assignments and also discuss the nonuniform
-back-off algorithms for the slotted media access design.
-
-% Start of "Sample References" section
-
 \section{Typical references in new ACM Reference Format}
@@
 
-% Appendix
-\appendix
-\section*{APPENDIX}
-\setcounter{section}{1}
-In this appendix, we measure
-the channel switching time of Micaz [CROSSBOW] sensor devices.
-In our experiments, one mote alternatingly switches between Channels
-11 and 12. Every time after the node switches to a channel, it sends
-out a packet immediately and then changes to a new channel as soon
-as the transmission is finished. We measure the
-number of packets the test mote can send in 10 seconds, denoted as
-$N_{1}$. In contrast, we also measure the same value of the test
-mote without switching channels, denoted as $N_{2}$. We calculate
-the channel-switching time $s$ as
-\begin{eqnarray}%
-s=\frac{10}{N_{1}}-\frac{10}{N_{2}}. \nonumber
-\end{eqnarray}%
-By repeating the experiments 100 times, we get the average
-channel-switching time of Micaz motes: 24.3$\mu$s.
+\section{Conclusions}
 
-\appendixhead{ZHOU}
+In this article, we develop the first multifrequency MAC protocol for
+WSN applications in which each device adopts a
+single radio transceiver. The different MAC design requirements for
+WSNs and general wireless ad-hoc networks are
+compared, and a complete WSN multifrequency MAC design (MMSN) is
+put forth. During the MMSN design, we analyze and evaluate different
+choices for frequency assignments and also discuss the nonuniform
+back-off algorithms for the slotted media access design.
 
@@
 \begin{acks}
-The authors would like to thank Dr. Maura Turolla of Telecom
-Italia for providing specifications about the application scenario.
+TODO The author would like to thank Dr. Penry for providing another great semester.
 \end{acks}
@@
 % History dates
-\received{February 2007}{March 2009}{June 2009}
-
-% Electronic Appendix
-\elecappendix
-
-\medskip
-
-\section{This is an example of Appendix section head}
-
-Channel-switching time is measured as the time length it takes for
-motes to successfully switch from one channel to another. This
-parameter impacts the maximum network throughput, because motes
-cannot receive or send any packet during this period of time, and it
-also affects the efficiency of toggle snooping in MMSN, where motes
-need to sense through channels rapidly.
-
-By repeating experiments 100 times, we get the average
-channel-switching time of Micaz motes: 24.3 $\mu$s. We then conduct
-the same experiments with different Micaz motes, as well as
-experiments with the transmitter switching from Channel 11 to other
-channels. In both scenarios, the channel-switching time does not have
-obvious changes. (In our experiments, all values are in the range of
-23.6 $\mu$s to 24.9 $\mu$s.)
-
-\section{Appendix section head}
-
-The primary consumer of energy in WSNs is idle listening. The key to
-reduce idle listening is executing low duty-cycle on nodes. Two
-primary approaches are considered in controlling duty-cycles in the
-MAC layer.
+\received{December 2015}{December 2015}{December 2015}
 
 \end{document}
-% End of v2-acmsmall-sample.tex (March 2012) - Gerry Murray, ACM

### 2015-12-05 FinalProject/Tex/paper/paper.tex
constructed preliminary outline for paper

@@
 \section{Introduction}
-
-As a new technology, Wireless Sensor Networks (WSNs) has a wide
-range of applications [Culler 2001,Bahl 2002,Akyildiz 2001], including
-environment monitoring, smart buildings, medical care, industrial and
-military applications. Among them, a recent trend is to develop
-commercial sensor networks that require pervasive sensing of both
-environment and human beings, for example, assisted living
-[Akyildiz 2002,Harvard 2001,CROSSBOW] and smart homes
-[Harvard 2001,Adya 2001,CROSSBOW].
-% quote
-\begin{quote}
-``For these applications, sensor devices are incorporated into human
-cloths [Natarajan 2001,Zhou 2006,Bahl 2002,Adya 2001] for monitoring
-health related information like EKG readings, fall detection, and voice recognition".
-\end{quote}
-While collecting all these multimedia information
-[Akyildiz 2002] requires a high network throughput, off-the-shelf
-sensor devices only provide very limited bandwidth in a single
-channel: 19.2Kbps in MICA2 [Bahl 2002] and 250Kbps in MICAz.
-
-In this article, we propose MMSN, abbreviation for Multifrequency
-Media access control for wireless Sensor Networks. The main
-contributions of this work can be summarized as follows.
-% itemize
 \begin{itemize}
-\item To the best of our knowledge, the MMSN protocol is the first
-multifrequency MAC protocol especially designed for WSNs, in which
-each device is equipped with a single radio transceiver and
-the MAC layer packet size is very small.
-\item Instead of using pairwise RTS/CTS frequency negotiation
-[Adya 2001,Culler 2001; Tzamaloukas 2001; Zhou 2006],
-we propose lightweight frequency assignments, which are good choices
-for many deployed comparatively static WSNs.
-\item We develop new toggle transmission and snooping techniques to
-enable a single radio transceiver in a sensor device to achieve
-scalable performance, avoiding the nonscalable ``one
-control channel + multiple data channels'' design [Natarajan 2001].
+\item State the purpose of the survey
+\item Explain why GPGPU is important and hard
+\item Explain what each of the following sections is
 \end{itemize}
 
-% Head 1
-\section{MMSN Protocol}
+\section{Background}
+\subsection{GPU History}
+TODO
+\subsection{GPGPUs and Current State}
+TODO
+\subsection{Applications}
+TODO
 
-% Head 2
-\subsection{Frequency Assignment}
+\section{Architectural Overview}
+\subsection{NVidia Warps}
+TODO
 
-We propose a suboptimal distribution to be used by each node, which is
-easy to compute and does not depend on the number of competing
-nodes. A natural candidate is an increasing geometric sequence, in
-which
-% Numbered Equation
-\begin{equation}
-\label{eqn:01}
-P(t)=\frac{b^{\frac{t+1}{T+1}}-b^{\frac{t}{T+1}}}{b-1},
-\end{equation}
-where $t=0,{\ldots}\,,T$, and $b$ is a number greater than $1$.
+\section{Challenges}
+\subsection{Heterogeneous Applications}
+TODO
+\subsection{Memory Limitations}
+TODO
 
-In our algorithm, we use the suboptimal approach for simplicity and
-generality. We need to make the distribution of the selected back-off
-time slice at each node conform to what is shown in Equation
-(\ref{eqn:01}). It is implemented as follows: First, a random
-variable $\alpha$ with a uniform distribution within the interval
-$(0, 1)$ is generated on each node, then time slice $i$ is selected
-according to the following equation:
-% Unnumbered Equation
-\[
-i=\lfloor(T+1)\log_b[\alpha(b-1)+1]\rfloor.
-\]
-It can be easily proven that the distribution of $i$ conforms to Equation
-(\ref{eqn:01}).
-
-So protocols [Bahl 2002,Culler 2001,Zhou 2006,Adya 2001,Culler 2001;
-Tzamaloukas-01; Akyildiz-01] that use RTS/CTS
-controls\footnote{RTS/CTS controls are required to be implemented by
-802.11-compliant devices. They can be used as an optional mechanism
-to avoid Hidden Terminal Problems in the 802.11 standard and
-protocols based on those similar to [Akyildiz 2001] and
-[Adya 2001].} for frequency negotiation and reservation are not
-suitable for WSN applications, even though they exhibit good
-performance in general wireless ad hoc
-networks.
-
-% Head 3
-\subsubsection{Exclusive Frequency Assignment}
-
-In exclusive frequency assignment, nodes first exchange their IDs
-among two communication hops so that each node knows its two-hop
-neighbors' IDs. In the second broadcast, each node beacons all
-neighbors' IDs it has collected during the first broadcast period.
-
-% Head 4
-\paragraph{Eavesdropping}
-
-Even though the even selection scheme leads to even sharing of
-available frequencies among any two-hop neighborhood, it involves a
-number of two-hop broadcasts. To reduce the communication cost, we
-propose a lightweight eavesdropping scheme.
-
-\subsection{Basic Notations}
-
-As Algorithm~\ref{alg:one} states, for each frequency
-number, each node calculates a random number (${\textit{Rnd}}_{\alpha}$) for
-itself and a random number (${\textit{Rnd}}_{\beta}$) for each of its two-hop
-neighbors with the same pseudorandom number generator.
-% Algorithm
-\begin{algorithm}[t]
-\SetAlgoNoLine
-\KwIn{Node $\alpha$'s ID ($ID_{\alpha}$), and node $\alpha$'s
-neighbors' IDs within two communication hops.}
-\KwOut{The frequency number ($FreNum_{\alpha}$) node $\alpha$ gets assigned.}
-$index$ = 0; $FreNum_{\alpha}$ = -1\;
-\Repeat{$FreNum_{\alpha} > -1$}{
-        $Rnd_{\alpha}$ = Random($ID_{\alpha}$, $index$)\;
-        $Found$ = $TRUE$\;
-        \For{each node $\beta$ in $\alpha$'s two communication hops
-    }{
-      $Rnd_{\beta}$ = Random($ID_{\beta}$, $index$)\;
-      \If{($Rnd_{\alpha} < Rnd_{\beta}$) \text{or} ($Rnd_{\alpha}$ ==
-          $Rnd_{\beta}$ \text{and} $ID_{\alpha} < ID_{\beta}$)\;
-      }{
-        $Found$ = $FALSE$; break\;
-      }
-        }
-     \eIf{$Found$}{
-           $FreNum_{\alpha}$ = $index$\;
-         }{
-           $index$ ++\;
-     }
-      }
-\caption{Frequency Number Computation}
-\label{alg:one}
-\end{algorithm}
-
-Bus masters are divided into two disjoint sets, $\mathcal{M}_{RT}$
-and $\mathcal{M}_{NRT}$.
-% description
-\begin{description}
-\item[RT Masters]
-$\mathcal{M}_{RT}=\{ \vec{m}_{1},\dots,\vec{m}_{n}\}$ denotes the
-$n$ RT masters issuing real-time constrained requests. To model the
-current request issued by an $\vec{m}_{i}$ in $\mathcal{M}_{RT}$,
-three parameters---the recurrence time $(r_i)$, the service cycle
-$(c_i)$, and the relative deadline $(d_i)$---are used, with their
-relationships.
-\item[NRT Masters]
-$\mathcal{M}_{NRT}=\{ \vec{m}_{n+1},\dots,\vec{m}_{n+m}\}$ is a set
-of $m$ masters issuing nonreal-time constrained requests. In our
-model, each $\vec{m}_{j}$ in $\mathcal{M}_{NRT}$ needs only one
-parameter, the service cycle, to model the current request it
-issues.
-\end{description}
-
-Here, a question may arise, since each node has a global ID. Why
-don't we just map nodes' IDs within two hops into a group of
-frequency numbers and assign those numbers to all nodes within two
-hops?
-
-\section{Simulator}
-\label{sec:sim}
-
-If the model checker requests successors of a state which are not
-created yet, the state space uses the simulator to create the
-successors on-the-fly. To create successor states the simulator
-conducts the following steps.
-% enumerate
-\begin{enumerate}
-\item Load state into microcontroller model.
-\item Determine assignments needed for resolving nondeterminism.
-\item For each assignment.
-      \begin{enumerate}
-      \item either call interrupt handler or simulate effect of next instruction, or
-      \item evaluate truth values of atomic propositions.
-      \end{enumerate}
-\item Return resulting states.
-\end{enumerate}
-Figure~\ref{fig:one} shows a typical microcontroller C program that
-controls an automotive power window lift. The program is one of the
-programs used in the case study described in Section~\ref{sec:sim}.
-At first sight, the programs looks like an ANSI~C program. It
-contains function calls, assignments, if clauses, and while loops.
-% Figure
-\begin{figure}
-% \centerline{\includegraphics{acmsmall-mouse}}
-\caption{Code before preprocessing.}
-\label{fig:one}
-\end{figure}
-
-\subsection{Problem Formulation}
-
-The objective of variable coalescence-based offset assignment is to find
-both the coalescence scheme and the MWPC on the coalesced graph. We start
-with a few definitions and lemmas for variable coalescence.
-
-% Enunciations
-\begin{definition}[Coalesced Node (C-Node)]A C-node is a set of
-live ranges (webs) in the AG or IG that are coalesced. Nodes within the same
-C-node cannot interfere with each other on the IG. Before any coalescing is
-done, each live range is a C-node by itself.
-\end{definition}
-
-\begin{definition}[C-AG (Coalesced Access Graph)]The C-AG is the access
-graph after node coalescence, which is composed of all C-nodes and C-edges.
-\end{definition}
-
-\begin{lemma}
-The C-MWPC problem is NP-complete.
-\end{lemma}
-\begin{proof} C-MWPC can be easily reduced to the MWPC problem assuming a
-coalescence graph without any edge or a fully connected interference graph.
-Therefore, each C-node is an uncoalesced live range after value separation
-and C-PC is equivalent to PC. A fully connected interference graph is made
-possible when all live ranges interfere with each other. Thus, the C-MWPC
-problem is NP-complete.
-\end{proof}
-
-\begin{lemma}[Lemma Subhead]The solution to the C-MWPC problem is no
-worse than the solution to the MWPC.
-\end{lemma}
-\begin{proof}
-Simply, any solution to the MWPC is also a solution to the
-C-MWPC. But some solutions to C-MWPC may not apply to the MWPC (if any
-coalescing were made).
-\end{proof}
-
-\section{Performance Evaluation}
-
-During all the experiments, the Geographic Forwarding (GF)
-[Akyildiz 2001] routing protocol is used. GF exploits geographic
-information of nodes and conducts local data-forwarding to achieve
-end-to-end routing. Our simulation is
-configured according to the settings in
-Table~\ref{tab:one}. Each run lasts for 2 minutes and
-repeated 100 times. For each data value we present in the results,
-we also give its 90\% confidence interval.
-% Table
-\begin{table}%
-\tbl{Simulation Configuration\label{tab:one}}{%
-\begin{tabular}{|l|l|}
-\hline
-TERRAIN{$^a$}   & (200m$\times$200m) Square\\\hline
-Node Number     & 289\\\hline
-Node Placement  & Uniform\\\hline
-Application     & Many-to-Many/Gossip CBR Streams\\\hline
-Payload Size    & 32 bytes\\\hline
-Routing Layer   & GF\\\hline
-MAC Layer       & CSMA/MMSN\\\hline
-Radio Layer     & RADIO-ACCNOISE\\\hline
-Radio Bandwidth & 250Kbps\\\hline
-Radio Range     & 20m--45m\\\hline
-\end{tabular}}
-\begin{tabnote}%
-\Note{Source:}{This is a table
-sourcenote. This is a table sourcenote. This is a table
-sourcenote.}
-\vskip2pt
-\Note{Note:}{This is a table footnote.}
-\tabnoteentry{$^a$}{This is a table footnote. This is a
-table footnote. This is a table footnote.}
-\end{tabnote}%
-\end{table}%
-
-\section{Typical references in new ACM Reference Format}
-A paginated journal article \cite{test1}, an enumerated
-journal article \cite{test1}, a reference to an entire issue \cite{test1},
-a monograph (whole book) \cite{test1}, a monograph/whole book in a series (see 2a in spec. document)
-\cite{test1}, a divisible-book such as an anthology or compilation \cite{test1}
-followed by the same example, however we only output the series if the volume number is given
-\cite{test1} (so Editor00a's series should NOT be present since it has no vol. no.),
-a chapter in a divisible book \cite{test1}, a chapter in a divisible book
-in a series \cite{test1}, a multi-volume work as book \cite{test1},
-an article in a proceedings (of a conference, symposium, workshop for example)
-(paginated proceedings article) \cite{test1}, a proceedings article
-with all possible elements \cite{test1}, an example of an enumerated
-proceedings article \cite{test1},
-an informally published work \cite{test1}, a doctoral dissertation \cite{test1},
-a master's thesis: \cite{test1}, an online document / world wide web resource \cite{test1}, \cite{test1},
-\cite{test1}, a video game (Case 1) \cite{test1} and (Case 2) \cite{test1}
-and \cite{test1} and (Case 3) a patent \cite{test1},
-work accepted for publication \cite{test1}, 'YYYYb'-test for prolific author
-\cite{test1} and \cite{test1}. Other cites might contain
-'duplicate' DOI and URLs (some SIAM articles) \cite{test1}.
-Boris / Barbara Beeton: multi-volume works as books
-\cite{test1} and \cite{test1}.
+\section{New Designs}
+TODO
 
 \section{Conclusions}
-
-In this article, we develop the first multifrequency MAC protocol for
-WSN applications in which each device adopts a
-single radio transceiver. The different MAC design requirements for
-WSNs and general wireless ad-hoc networks are
-compared, and a complete WSN multifrequency MAC design (MMSN) is
-put forth. During the MMSN design, we analyze and evaluate different
-choices for frequency assignments and also discuss the nonuniform
-back-off algorithms for the slotted media access design.
+\begin{itemize}
+\item Mention what the survey talked about
+  \begin{itemize}
+  \item Background
+  \item Overview
+  \item Challenges
+  \item Designs
+  \end{itemize}
+\item Explain changes occurring in field
+\item Talk about what the future could hold
+\end{itemize}
 
@@
 \begin{acks}
-TODO The author would like to thank Dr. Penry for providing another great semester.
+TODO The author would like to thank Dr. Penry for teaching this course and providing extensive design and learning opportunities.
 \end{acks}
 
-% Bibliography
-\bibliographystyle{ACM-Reference-Format-Journals}
-\bibliography{paper-bibfile}
-                             % Sample .bib file with references that match those in
-                             % the 'Specifications Document (V1.5)' as well containing
-                             % 'legacy' bibs and bibs with 'alternate codings'.
-                             % Gerry Murray - March 2012
+% % Bibliography
+% \bibliographystyle{ACM-Reference-Format-Journals}
+% \bibliography{paper-bibfile}
 

### 2015-12-07 FinalProject/Tex/paper/paper.tex
started outlining paper

@@
 TODO
+Multiple threads are grouped together in "warps"
+Thread within a warp are mapped to a vector functional unit (SIMD)
+NVIDIA's Maxwell GPUs support up to 64 concurrent warps (2048 threads)
+Talk about what a streaming multiprocessor is
+Modern unified GPUs consist of multiple streaming multiprocessors (SMs)
+  - warp (NVIDIA)/wavefronts (OpenCL)
+  - similar to a SIMD processor
+  - SM has a warp pool to store context of all running threads
+  - HW warp scheduler dispatches and allocates resources
+  - large (much larger than a CPU) register file shared across all threads
+    - Maxwell architecture has 64K 32-bit registers for each SM
+  - kernel: large number of threads executing same piece of code
+  - multiple threads form a thread-block; same life-cycle, dispatched together
+  - threads within a thread-block (32 or 64) organized into warps
+  - warp scheduler schedules a ready warp each cycle, hiding latency of stalled warps
+  - CPU launches GPU kernels by dispatching launch commands. Kernel parameters are passed from GPU to CPU at launch time and stored in GPU global memory
+  - kernels from different streams are independent and can be run concurrently, but kernels within a stream are sequential
+  - see DTBL paper for good intro
+  - also see Locality Aware paper for internals of a SM
+  - cooperative thread array (CTA) - see Weak behavior paper
 
@@
 TODO
+Maximizing resource utilization while minimizing shared resource contention between CPU and GPU is hard
 \subsection{Memory Limitations}
@@
 \section{New Designs}
-TODO
+Warp compression stuff
+  Assist warps (CABA)
+  Warped Compression
+Dynamic tuning
+  Coordinated warp scheduling (CAWA)
+  Dynamic Thread Block Launch (DTBL)
+  Equalizer
+  Fine-grain Hardware Worklists (HWWL)
+  Variable Warp Size (VWS)
+CPU/GPU contention, NoC
+  Managing Concurrency in Heterogeneous
+Compiler/SW-only design
+  Divergence Management
+  Locality Aware Mapping
+  (x) WeakBehaviorProgrammingAssumptions
 

### 2015-12-07 FinalProject/Tex/paper/paper.tex
added title, fake abstract, keywords, bulleted ideas

@@
 
+\newcommand{\titleName}{General-Purpose GPUs and the Challenges of Efficient
+Parallel Processing of Heterogeneous Applications}
+
 % Page heads
-\markboth{C. Smith}{TODO A Survey Paper}
+\markboth{C. Smith}{\titleName{}}
 
 % Title portion
-\title{TODO A Survey Paper}
+\title{\titleName{}}
 \author{Connor Smith
@@
 \begin{abstract}
-This is my abstract.
+TODO This is my abstract. It has many pretty lorem ipsum words. Lorem ipsum
+dolor sit amet, consectetur adipiscing elit. Aliquam nec est at nibh tempus
+malesuada. Proin eget risus ut arcu imperdiet porttitor nec maximus est. Ut
+vulputate gravida arcu vitae molestie. Vivamus tempor semper erat in
+condimentum. Vestibulum tincidunt pulvinar dolor, sit amet tincidunt sem
+sollicitudin et. Donec ut ultricies nulla, fringilla gravida neque. Sed feugiat
+mauris eu enim ultricies, id tincidunt velit semper. Curabitur blandit diam
+odio, in mollis risus porttitor eget. Vivamus rutrum enim turpis, eu posuere
+nibh mattis porttitor. Aenean mollis augue nec semper porta. Nulla semper vitae
+sapien quis commodo. Mauris tempor libero sollicitudin cursus lacinia.
 \end{abstract}
 
-\keywords{TODO, General purpose graphics processing unit, heterogeneous systems,
-warp size}
+\keywords{General purpose graphics processing unit, heterogeneous systems, warp
+size, dynamic warp, GPU compression, hardware worklists, locality aware mapping}
 
-\acmformat{Connor Smith, 2015. TODO A Survey Paper.}
+\acmformat{\titleName{}.}
 
 \begin{bottomstuff}
-This work is supported by the National Imaginary Science Foundation, under
-grant CNS-0123456.
+This work is supported by the National Imaginary Science Foundation, under grant
+CNS-0123456.
 
-Author's addresses: C. Smith, Electrical and Computer Engineering Department,
+Author's address: C. Smith, Electrical and Computer Engineering Department,
 Brigham Young University.
@@
 \section{New Designs}
-Warp compression stuff
-  Assist warps (CABA)
-  Warped Compression
-Dynamic tuning
-  Coordinated warp scheduling (CAWA)
-  Dynamic Thread Block Launch (DTBL)
-  Equalizer
-  Fine-grain Hardware Worklists (HWWL)
-  Variable Warp Size (VWS)
-CPU/GPU contention, NoC
-  Managing Concurrency in Heterogeneous
-Compiler/SW-only design
-  Divergence Management
-  Locality Aware Mapping
-  (x) WeakBehaviorProgrammingAssumptions
+\begin{itemize}
+\item Warp compression stuff
+  \begin{itemize}
+  \item Assist warps (CABA)
+  \item Warped Compression
+  \end{itemize}
+\item Dynamic tuning
+  \begin{itemize}
+  \item Coordinated warp scheduling (CAWA)
+  \item Dynamic Thread Block Launch (DTBL)
+  \item Equalizer
+  \item Fine-grain Hardware Worklists (HWWL)
+  \item Variable Warp Size (VWS)
+  \end{itemize}
+\item CPU/GPU contention, NoC
+  \begin{itemize}
+  \item Managing Concurrency in Heterogeneous
+  \end{itemize}
+\item Compiler/SW-only design
+  \begin{itemize}
+  \item Divergence Management
+  \item Locality Aware Mapping
+  \item (x) WeakBehaviorProgrammingAssumptions
+  \end{itemize}
+\end{itemize}
 
@@
 \begin{acks}
-TODO The author would like to thank Dr. Penry for teaching this course and providing extensive design and learning opportunities.
+TODO The author would like to thank Dr. Penry for teaching this course and
+providing extensive design and learning opportunities.
 \end{acks}

### 2015-12-07 FinalProject/Tex/paper/paper.tex
added abstract

@@
 \begin{abstract}
-TODO This is my abstract. It has many pretty lorem ipsum words. Lorem ipsum
-dolor sit amet, consectetur adipiscing elit. Aliquam nec est at nibh tempus
-malesuada. Proin eget risus ut arcu imperdiet porttitor nec maximus est. Ut
-vulputate gravida arcu vitae molestie. Vivamus tempor semper erat in
-condimentum. Vestibulum tincidunt pulvinar dolor, sit amet tincidunt sem
-sollicitudin et. Donec ut ultricies nulla, fringilla gravida neque. Sed feugiat
-mauris eu enim ultricies, id tincidunt velit semper. Curabitur blandit diam
-odio, in mollis risus porttitor eget. Vivamus rutrum enim turpis, eu posuere
-nibh mattis porttitor. Aenean mollis augue nec semper porta. Nulla semper vitae
-sapien quis commodo. Mauris tempor libero sollicitudin cursus lacinia.
+This article is a survey of the current state of general-purpose computing on
+GPUs (GPGPUs) and the use of heterogeneous applications on them. The underlying
+architecture of GPUs is well-suited to problems with a high level of data
+parallelism and has traditionally been used for rendering computer graphics.
+Performing general-purpose computing on GPUs is a relatively recent development,
+having only become practical and popular in the past decade or so. A brief
+overview of a typical GPU architecture is presented, and the challenges of
+applying this architecture to heterogeneous problems is discussed. Several
+different current areas of research are explored, including warp compression,
+dynamic tuning of warps, managing contention between CPUs and GPUs, and compiler
+management of heterogeneous applications. While the challenges presented are
+difficult to overcome, general-purpose computing on GPUs continues to make
+progress in the areas of performance, energy, and power.
 \end{abstract}

### 2015-12-07 FinalProject/Tex/paper/paper.tex
removed extraneous comment
added introduction

@@
 
-% % Metadata Information
+% Metadata Information
 \acmVolume{000}
@@
 \section{Introduction}
-\begin{itemize}
-\item State the purpose of the survey
-\item Explain why GPGPU is important and hard
-\item Explain what each of the following sections is
-\end{itemize}
+The purpose of this survey is to provide an introduction to general-purpose
+computing on graphics processing units (GPGPU), the challenges inherent to this
+field, and some active areas of research within the field to alleviate some of
+these problems. To the best of the author's knowledge, this is the first survey
+in this semester's course to discuss GPGPU.
 
-\section{Background}
-\subsection{GPU History}
-TODO
-\subsection{GPGPUs and Current State}
-TODO
-\subsection{Applications}
-TODO
+The term GPU was popularized around the turn of the 21st century by NVIDIA as
+consumer-level devices began to need dedicated accelerators to render 3D
+graphics. Today, most all computers contain GPUs, from supercomputers to
+laptops, desktops to smartphones, although some of these do not use dedicated
+graphics cards. The area of GPGPU started to take off roughly a decade ago when it began to be
+practical to solve traditional problems with the massively parallel nature of
+GPUs. Using hundreds or even thousands of ``cores" to perform computations in
+parallel is tempting (even if the cores aren't really cores at all), but brings
+with it a new set of challenges.
 
-\section{Architectural Overview}
+In a way, the problems of GPGPU are an extension of those sometimes seen in a
+traditional desktop multiprocessing system. Even when under a heavy compute
+load, it isn't always easy to keep all of the cores doing useful work in any
+given cycle. Between limitations in structural units, dependences between
+instructions, and latencies of memory accesses, the processor cannot help but
+stall sometimes. This problem is exacerbated by orders of magnitude when solving
+heterogeneous problems on a GPU. GPUs were originally designed to operate on
+graphics processing, problems that are well-structured and have simple, inherent
+data parallelism. Problems that do not contain parallelism that is easily mapped
+to the architecture of a GPU fail to keep all of the cores in a GPU busy or
+strain the memory bandwidth past the GPU's ability to hide latency.
+
+The remainder of this survey is organized as follows. Section
+\ref{sec:architecture} provides a brief architectural overview of GPUs. Section
+\ref{sec:challenges} elaborates on the challenges specific to GPGPU
+applications. Section \ref{sec:research} summarizes and discusses several
+current areas of research related to GPGPU. Section \ref{sec:conclusions}
+concludes the survey.
+
+% \section{Background}
+% \subsection{GPU History}
+% TODO
+% \subsection{GPGPUs and Current State}
+% TODO
+% \subsection{Applications}
+% TODO
+
+\section{Architectural Overview} \label{sec:architecture}
 \subsection{NVidia Warps}
@@
 
-\section{Challenges}
+\section{Challenges} \label{sec:challenges}
 \subsection{Heterogeneous Applications}
@@
 
-\section{New Designs}
+\section{New Designs} \label{sec:research}
 \begin{itemize}
@@
 
-\section{Conclusions}
+\section{Conclusions} \label{sec:conclusions}
 \begin{itemize}

### 2015-12-07 FinalProject/Tex/paper/paper.tex
started working on architecture section

@@
 \section{Architectural Overview} \label{sec:architecture}
-\subsection{NVidia Warps}
-TODO
-Multiple threads are grouped together in "warps"
-Thread within a warp are mapped to a vector functional unit (SIMD)
-NVIDIA's Maxwell GPUs support up to 64 concurrent warps (2048 threads)
-Talk about what a streaming multiprocessor is
+\subsection{Streaming Multiprocessors}
 Modern unified GPUs consist of multiple streaming multiprocessors (SMs)
-  - warp (NVIDIA)/wavefronts (OpenCL)
-  - similar to a SIMD processor
-  - SM has a warp pool to store context of all running threads
-  - HW warp scheduler dispatches and allocates resources
-  - large (much larger than a CPU) register file shared across all threads
-    - Maxwell architecture has 64K 32-bit registers for each SM
-  - kernel: large number of threads executing same piece of code
-  - multiple threads form a thread-block; same life-cycle, dispatched together
-  - threads within a thread-block (32 or 64) organized into warps
-  - warp scheduler schedules a ready warp each cycle, hiding latency of stalled warps
-  - CPU launches GPU kernels by dispatching launch commands. Kernel parameters are passed from GPU to CPU at launch time and stored in GPU global memory
-  - kernels from different streams are independent and can be run concurrently, but kernels within a stream are sequential
-  - see DTBL paper for good intro
-  - also see Locality Aware paper for internals of a SM
-  - cooperative thread array (CTA) - see Weak behavior paper
+\begin{itemize}
+  \item warp (NVIDIA)/wavefronts (OpenCL)
+  \item similar to a SIMD processor
+  \item SM has a warp pool to store context of all running threads
+  \item HW warp scheduler dispatches and allocates resources
+  \item large (much larger than a CPU) register file shared across all threads
+    \item Maxwell architecture has 64K 32-bit registers for each SM
+  \item kernel: large number of threads executing same piece of code
+  \item multiple threads form a thread-block; same life-cycle, dispatched together
+  \item threads within a thread-block (32 or 64) organized into warps
+  \item warp scheduler schedules a ready warp each cycle, hiding latency of stalled warps
+  \item CPU launches GPU kernels by dispatching launch commands. Kernel parameters are passed from GPU to CPU at launch time and stored in GPU global memory
+  \item kernels from different streams are independent and can be run concurrently, but kernels within a stream are sequential
+  \item see DTBL paper for good intro
+  \item also see Locality Aware paper for internals of a SM
+  \item cooperative thread array (CTA) - see Weak behavior paper
+\end{itemize}
+\subsection{NVIDIA Warps}
+\begin{itemize}
+  \item Multiple threads are grouped together in ``warps''
+  \item Thread within a warp are mapped to a vector functional unit (SIMD)
+  \item NVIDIA's Maxwell GPUs support up to 64 concurrent warps (2048 threads)
+\end{itemize}
 
-\section{Challenges} \label{sec:challenges}
+\section{Challenges of GPGPU} \label{sec:challenges}
 \subsection{Heterogeneous Applications}

### 2015-12-07 FinalProject/Tex/paper/paper.tex
shortened title across the top of the page

@@
 
-\newcommand{\titleName}{General-Purpose GPUs and the Challenges of Efficient
+\newcommand{\titleNameLong}{General-Purpose GPUs and the Challenges of Efficient
+Parallel Processing of Heterogeneous Applications}
+
+\newcommand{\titleNameShort}{GPGPU and the Challenges of Efficient
 Parallel Processing of Heterogeneous Applications}
@@
 % Page heads
-\markboth{C. Smith}{\titleName{}}
+\markboth{C. Smith}{\titleNameShort{}}
 
 % Title portion
-\title{\titleName{}}
+\title{\titleNameLong{}}
 \author{Connor Smith
@@
 
-\acmformat{\titleName{}.}
+\acmformat{\titleNameLong{}.}
 

### 2015-12-07 FinalProject/Tex/paper/paper.tex
updated acknowledgements

@@
 \begin{acks}
-TODO The author would like to thank Dr. Penry for teaching this course and
-providing extensive design and learning opportunities.
+The author would like to thank Dr. Penry for teaching this course and for
+providing extensive design and learning opportunities in this course. It has
+been a lot of work, but this course has helped the author know that he has
+chosen the right field.
 \end{acks}

### 2015-12-07 FinalProject/Tex/paper/paper.tex
worked on architecture section

@@
 \subsection{Streaming Multiprocessors}
-Modern unified GPUs consist of multiple streaming multiprocessors (SMs)
 \begin{itemize}
-  \item warp (NVIDIA)/wavefronts (OpenCL)
+  \item Modern unified GPUs consist of multiple streaming multiprocessors (SMs)
   \item similar to a SIMD processor
-  \item SM has a warp pool to store context of all running threads
-  \item HW warp scheduler dispatches and allocates resources
-  \item large (much larger than a CPU) register file shared across all threads
-    \item Maxwell architecture has 64K 32-bit registers for each SM
-  \item kernel: large number of threads executing same piece of code
-  \item multiple threads form a thread-block; same life-cycle, dispatched together
-  \item threads within a thread-block (32 or 64) organized into warps
-  \item warp scheduler schedules a ready warp each cycle, hiding latency of stalled warps
-  \item CPU launches GPU kernels by dispatching launch commands. Kernel parameters are passed from GPU to CPU at launch time and stored in GPU global memory
-  \item kernels from different streams are independent and can be run concurrently, but kernels within a stream are sequential
-  \item see DTBL paper for good intro
-  \item also see Locality Aware paper for internals of a SM
-  \item cooperative thread array (CTA) - see Weak behavior paper
+  \item A SM contains
+  \begin{itemize}
+    \item a large register file shared across all threads. This register file
+    typically contains thousands of registers (the NVIDIA Maxwell architecture
+    has 64K 32-bit registers in each SM)
+    \item multiple caches. Maxwell includes a unified L1/texture cache used as a
+    coalescing buffer for memory accesses. The buffer gathers requested data and
+    sends the entire group to a warp.
+    \item shared memory (96KB per SM). Thread blocks are limited to 48KB each.
+    Shared memory for a thread-block is only accessible by threads within the
+    same thread-block.
+    \item multiple warp schedulers. A warp scheduler dispatches and allocates
+    resources. Maxwell has four per SM. FUs are assigned to a particular
+    scheduler. Power of two CUDA cores per SM allows for simple partitioning
+    among schedulers. One ready warp is scheduled per cycle, hiding the latency
+    of stalled warps.
   \end{itemize}
-\subsection{NVIDIA Warps}
+
+\end{itemize}
+\subsection{Kernels, Thread-blocks, and Warps}
 \begin{itemize}
-  \item Multiple threads are grouped together in ``warps''
-  \item Thread within a warp are mapped to a vector functional unit (SIMD)
-  \item NVIDIA's Maxwell GPUs support up to 64 concurrent warps (2048 threads)
+  \item A kernel is a massive number of threads executing same piece of code.
+  Kernels from different streams are independent and can be run concurrently,
+  but kernels within a stream are sequential.
+  \item The CPU launches GPU kernels by dispatching launch commands. Kernel
+  parameters are passed from the GPU to CPU at launch time and are stored in the
+  GPU's global memory.
+  \item Grids are what actually run on the GPU and are composed of multiple
+  ``thread- blocks'', also called a cooperative thread array (CTA). Thread-
+  blocks contain on the order of 512 threads. Each thread within a thread-block
+  has the same life-cycle as all the others, meaning that they are dispatched to
+  and swapped from a SM at the same time.
+  \item Individual threads exist within a thread-block. These threads are
+  grouped, usually around 32 at a time, into ``warps''. The warp is the
+  fundamental unit of execution on GPUs. Warp is NVIDIA's term; AMD and OpenCL
+  use the term ``wavefront''. Maxwell supports up to 64 warps resident per SM.
+  This doesn't mean that all 64 are executing at the same time, but that the
+  architecture can support that many warp contexts concurrently.
+  \item The SM maintains a warp pool to store the context of all running
+  threads. Warps are scheduled by the warp schedulers from this pool.
+  \item Threads within a warp are mapped to a vector functional unit such that
+  all threads execute the same instruction on different data, just a like a SIMD
+  unit.
 \end{itemize}

### 2015-12-07 FinalProject/Tex/paper/paper.tex
fixed error in architecture section
fixed word wrapping issue
converted bulleted list to paragraphs

@@
 laptops, desktops to smartphones, although some of these do not use dedicated
-graphics cards. The area of GPGPU started to take off roughly a decade ago when it began to be
-practical to solve traditional problems with the massively parallel nature of
-GPUs. Using hundreds or even thousands of ``cores" to perform computations in
-parallel is tempting (even if the cores aren't really cores at all), but brings
-with it a new set of challenges.
+graphics cards. The area of GPGPU started to take off roughly a decade ago when
+it began to be practical to solve traditional problems with the massively
+parallel nature of GPUs. Using hundreds or even thousands of ``cores" to perform
+computations in parallel is tempting (even if the cores aren't really cores at
+all), but brings with it a new set of challenges.
 
@@
 \subsection{Streaming Multiprocessors}
-\begin{itemize}
-  \item Modern unified GPUs consist of multiple streaming multiprocessors (SMs)
-  \item similar to a SIMD processor
-  \item A SM contains
-  \begin{itemize}
-    \item a large register file shared across all threads. This register file
-    typically contains thousands of registers (the NVIDIA Maxwell architecture
-    has 64K 32-bit registers in each SM)
-    \item multiple caches. Maxwell includes a unified L1/texture cache used as a
-    coalescing buffer for memory accesses. The buffer gathers requested data and
-    sends the entire group to a warp.
-    \item shared memory (96KB per SM). Thread blocks are limited to 48KB each.
-    Shared memory for a thread-block is only accessible by threads within the
-    same thread-block.
-    \item multiple warp schedulers. A warp scheduler dispatches and allocates
-    resources. Maxwell has four per SM. FUs are assigned to a particular
-    scheduler. Power of two CUDA cores per SM allows for simple partitioning
-    among schedulers. One ready warp is scheduled per cycle, hiding the latency
-    of stalled warps.
-  \end{itemize}
+Modern unified GPUs consist of multiple streaming multiprocessors (SMs). While
+there are many GPU architectures available today, the following description is
+based off of NVIDIA's Maxwell architecture, which was released in 2014. These
+SMs are similar to SIMD processors and typically contain the following
+components:
+\begin{description}
+  \setlength\itemsep{1em}
+  \item[A large register file] The register file is shared across all threads
+  and typically contains thousands of registers, many more than are usually
+  found in a CPU. NVIDIA's Maxwell architecture has 64K 32-bit registers in each
+  SM.
+  \item[Multiple caches] Maxwell includes a unified L1/texture cache, which used
+  as a coalescing buffer for memory accesses. The buffer gathers requested data
+  and sends the entire group to a warp. There are also additional caches TODO.
+  \item[Shared memory] There are 96KB available per SM in Maxwell. However, thread blocks
+  are limited to 48KB each. Shared memory for a thread-block is only accessible
+  by threads within the same thread-block.
+  \item[Multiple warp schedulers] Warp schedulers dispatch and allocate
+  resources. Maxwell contains four schedulers per SM, and functional units are
+  assigned to a particular scheduler. Using a power-of-two number of CUDA cores
+  per SM allows for simple partitioning among schedulers. One ready warp is
+  scheduled per cycle, which allows the GPU to hide the latency of stalled
+  warps.
+\end{description}
 
-\end{itemize}
 \subsection{Kernels, Thread-blocks, and Warps}
-\begin{itemize}
-  \item A kernel is a massive number of threads executing same piece of code.
-  Kernels from different streams are independent and can be run concurrently,
-  but kernels within a stream are sequential.
-  \item The CPU launches GPU kernels by dispatching launch commands. Kernel
-  parameters are passed from the GPU to CPU at launch time and are stored in the
-  GPU's global memory.
-  \item Grids are what actually run on the GPU and are composed of multiple
-  ``thread- blocks'', also called a cooperative thread array (CTA). Thread-
-  blocks contain on the order of 512 threads. Each thread within a thread-block
-  has the same life-cycle as all the others, meaning that they are dispatched to
-  and swapped from a SM at the same time.
-  \item Individual threads exist within a thread-block. These threads are
-  grouped, usually around 32 at a time, into ``warps''. The warp is the
-  fundamental unit of execution on GPUs. Warp is NVIDIA's term; AMD and OpenCL
-  use the term ``wavefront''. Maxwell supports up to 64 warps resident per SM.
-  This doesn't mean that all 64 are executing at the same time, but that the
-  architecture can support that many warp contexts concurrently.
-  \item The SM maintains a warp pool to store the context of all running
-  threads. Warps are scheduled by the warp schedulers from this pool.
-  \item Threads within a warp are mapped to a vector functional unit such that
-  all threads execute the same instruction on different data, just a like a SIMD
-  unit.
-\end{itemize}
+A kernel is a potentially massive number of threads executing same piece of
+code. From the application programmer's point of view, a function is called on
+the device with a certain number of blocks and a certain number of threads
+within each block. The CPU launches GPU kernels by dispatching launch commands.
+Kernel parameters are passed from the CPU to the GPU at launch time and are
+stored in the GPU's global memory. Kernels from different streams are
+independent and can be run concurrently, but kernels within a stream are
+sequential.
+
+Grids are what actually run on the GPU, and they are composed of multiple
+``thread- blocks'', also called a cooperative thread array (CTA) outside of
+NVIDIA's nomenclature. Maxwell supports up to 64K blocks within a grid. Thread-
+blocks contain potentially many threads, on the order of 1024 for modern
+architectures. Each thread within a thread-block has the same life-cycle as all
+the others, meaning that they are dispatched to and swapped from a SM at the
+same time.
+
+Individual threads exist within a thread-block. These threads are grouped into
+``warps'', usually around 32 at a time. The warp is the fundamental unit of
+execution on GPUs. Threads within a warp are mapped to a vector functional unit
+such that all threads execute the same instruction on different data, just a
+like a SIMD unit. With this in mind, it would be more accurate to simply call
+``threads'' a sequence of SIMD lane instructions, but this survey will continue
+to refer to them as threads to maintain consistency with the literature. The
+term ``Warp'' is the term used by NVIDIA, while AMD and OpenCL use
+``wavefront''. Maxwell supports up to 64 warps resident per SM. Note that this
+doesn't mean that all 64 warps are executing at the same time, but that the
+architecture can support that many warp contexts concurrently. The SM maintains
+a warp pool to store the context of all running threads. Warps are scheduled by
+the warp schedulers from this pool.
 
@@
 \subsection{Heterogeneous Applications}
-TODO
-Maximizing resource utilization while minimizing shared resource contention between CPU and GPU is hard
+TODO Maximizing resource utilization while minimizing shared resource contention
+between CPU and GPU is hard
 \subsection{Memory Limitations}

### 2015-12-07 FinalProject/Tex/paper/paper.tex
added bibliography info
fixed another word wrap issue
added challenges section

@@
   and sends the entire group to a warp. There are also additional caches TODO.
-  \item[Shared memory] There are 96KB available per SM in Maxwell. However, thread blocks
-  are limited to 48KB each. Shared memory for a thread-block is only accessible
-  by threads within the same thread-block.
+  \item[Shared memory] There are 96KB available per SM in Maxwell. However,
+  thread blocks are limited to 48KB each. Shared memory for a thread-block is
+  only accessible by threads within the same thread-block.
   \item[Multiple warp schedulers] Warp schedulers dispatch and allocate
@@
 \subsection{Heterogeneous Applications}
-TODO Maximizing resource utilization while minimizing shared resource contention
-between CPU and GPU is hard
+There are two main challenges of GPGPU. The first is maintaining high
+performance on heterogeneous applications. As mentioned previously, this problem
+is difficult simply due to the complicated structure of these applications. When
+a warp stalls for one reason or another, the warp scheduler selects a new warp
+to take its place the next cycle. This is an attempt to hide the latency of warp
+stalls. However, GPU schedulers are limited in their ability to hide latencies
+for heterogeneous applications. This is caused by imbalanced workloads among
+warps, divergent branches, irregular memory accesses, and shared cache
+contention \cite{CoordinatedWarpScheduling}. Designs in the following section
+discusses these issues and present possible solutions.
+
 \subsection{Memory Limitations}
-TODO
+The second challenge for GPGPU is mitigating memory bandwidth bottlenecks. A
+majority of general-purpose applications run on GPUs are memory bound
+\cite{AssistWarps}, which means that performance is bottlenecked by off-chip
+memory bandwidth. This bottleneck is commonly due to contention among warps
+within the GPU, and may also be a result of contention with memory use by the
+CPU in heterogeneous system architectures
+\cite{ManagingConcurrencyInHeterogeneous}. A number of schemes have been
+suggested to reduce pressure on off-chip memory and will be visited in the next
+section.
 
@@
 
-% % Bibliography
-% \bibliographystyle{ACM-Reference-Format-Journals}
-% \bibliography{paper-bibfile}
+% Bibliography
+\bibliographystyle{ACM-Reference-Format-Journals}
+\bibliography{paper-bibfile}
 

### 2015-12-07 FinalProject/Tex/paper/paper.tex
tweaked title

@@
 
-\newcommand{\titleNameLong}{General-Purpose GPUs and the Challenges of Efficient
-Parallel Processing of Heterogeneous Applications}
+\newcommand{\titleNameLong}{General-Purpose Computing on GPUs and Efficient Parallel
+Processing of Heterogeneous Applications}
 
-\newcommand{\titleNameShort}{GPGPU and the Challenges of Efficient
-Parallel Processing of Heterogeneous Applications}
+\newcommand{\titleNameShort}{GPGPU and Efficient Parallel Processing of
+Heterogeneous Applications}
 

### 2015-12-07 FinalProject/Tex/paper/paper.tex
converted design section from list to subsections
expanded subsections into sections

@@
 \ref{sec:challenges} elaborates on the challenges specific to GPGPU
-applications. Section \ref{sec:research} summarizes and discusses several
-current areas of research related to GPGPU. Section \ref{sec:conclusions}
-concludes the survey.
-
-% \section{Background}
-% \subsection{GPU History}
-% TODO
-% \subsection{GPGPUs and Current State}
-% TODO
-% \subsection{Applications}
-% TODO
+applications. Section \ref{sec:dynamic} discusses various dynamic approaches to
+scheduling and warp configurations. Section \ref{sec:memory} talks specifically
+about reducing memory contention. Section \ref{sec:software} is a discussion of
+a few software and compiler solutions to the challenges of GPGPU. Section
+\ref{sec:conclusions} concludes the survey.
 
@@
 
-\section{New Designs} \label{sec:research}
-\begin{itemize}
-\item Warp compression stuff
-  \begin{itemize}
-  \item Assist warps (CABA)
-  \item Warped Compression
-  \end{itemize}
-\item Dynamic tuning
-  \begin{itemize}
-  \item Coordinated warp scheduling (CAWA)
-  \item Dynamic Thread Block Launch (DTBL)
-  \item Equalizer
-  \item Fine-grain Hardware Worklists (HWWL)
-  \item Variable Warp Size (VWS)
-  \end{itemize}
-\item CPU/GPU contention, NoC
-  \begin{itemize}
-  \item Managing Concurrency in Heterogeneous
-  \end{itemize}
-\item Compiler/SW-only design
-  \begin{itemize}
-  \item Divergence Management
-  \item Locality Aware Mapping
-  \item (x) WeakBehaviorProgrammingAssumptions
-  \end{itemize}
-\end{itemize}
+\section{Dynamic Warp Scheduling and Hardware Tuning} \label{sec:dynamic}
+\subsection{Assist warps (CABA)}
+This section left as an exercise to the reader
+\subsection{Coordinated warp scheduling (CAWA)}
+This section left as an exercise to the reader
+\subsection{Dynamic Thread Block Launch (DTBL)}
+This section left as an exercise to the reader
+\subsection{Equalizer}
+This section left as an exercise to the reader
+\subsection{Fine-grain Hardware Worklists (HWWL)}
+This section left as an exercise to the reader
+\subsection{Variable Warp Size (VWS)}
+This section left as an exercise to the reader
+
+\section{Reducing Memory Contention} \label{sec:memory}
+\subsection{Warped Compression}
+This section left as an exercise to the reader
+\subsection{Managing Concurrency in Heterogeneous (CPU/GPU NoC contention)}
+This section left as an exercise to the reader
+
+\section{Compiler/SW-only design} \label{sec:software}
+\subsection{Divergence Management}
+This section left as an exercise to the reader
+\subsection{Locality Aware Mapping}
+This section left as an exercise to the reader
+\subsection{(x) WeakBehaviorProgrammingAssumptions}
+This section left as an exercise to the reader
 

### 2015-12-07 FinalProject/Tex/paper/paper.tex
fixed grammatical issue
reworded some of the abstract

@@
 overview of a typical GPU architecture is presented, and the challenges of
-applying this architecture to heterogeneous problems is discussed. Several
-different current areas of research are explored, including warp compression,
-dynamic tuning of warps, managing contention between CPUs and GPUs, and compiler
-management of heterogeneous applications. While the challenges presented are
-difficult to overcome, general-purpose computing on GPUs continues to make
-progress in the areas of performance, energy, and power.
+applying this architecture to heterogeneous problems are discussed. Several
+different current areas of research are explored, dynamic warp scheduling and
+hardware tuning, reducing memory contention, and software solutions to the
+challenges of GPGPU. While the challenges presented are difficult to overcome,
+general-purpose computing on GPUs continues to make progress in the areas of
+performance, energy, and power.
 \end{abstract}

### 2015-12-08 FinalProject/Tex/paper/paper.tex
fixed line breaks on title
split dynamic section in two, finished first half

@@
 
-\newcommand{\titleNameLong}{General-Purpose Computing on GPUs and Efficient Parallel
-Processing of Heterogeneous Applications}
+\newcommand{\titleNameLong}{General-Purpose Computing on GPUs and Efficient
+Parallel Processing of Heterogeneous Applications}
 
@@
 \ref{sec:challenges} elaborates on the challenges specific to GPGPU
-applications. Section \ref{sec:dynamic} discusses various dynamic approaches to
-scheduling and warp configurations. Section \ref{sec:memory} talks specifically
-about reducing memory contention. Section \ref{sec:software} is a discussion of
-a few software and compiler solutions to the challenges of GPGPU. Section
+applications. Section \ref{sec:dynamic} discusses spawning granular helpers to
+increase performance. Section \ref{sec:scheduling} deals with more efficient
+scheduling within and among streaming multiprocessors. Section \ref{sec:memory}
+talks specifically about reducing memory contention. Section \ref{sec:software}
+is a discussion of a few software and compiler solutions to the challenges of
+GPGPU. Section
 \ref{sec:conclusions} concludes the survey.
@@
 
-\section{Dynamic Warp Scheduling and Hardware Tuning} \label{sec:dynamic}
-\subsection{Assist warps (CABA)}
+\section{Dynamically-Launched Helpers} \label{sec:dynamic}
+\subsection{Assist warps}
+As we have seen, GPUs are capable of supporting thousands of threads, but
+bottlenecks in heterogeneous applications limit performance. \cite{AssistWarps}
+present the idea of using ``assist warps'' in their Core-Assisted Bottleneck
+Acceleration (CABA) scheme. These special warps are generated by dedicated
+hardware and are used as helper threads to increase performance and efficiency.
+Assist warps share the same context as the parent warp and uses some of the
+available registers, since investigation shows that 24\% of the register file is
+unallocated on average.
+
+One primary use for assist warps is memory compression. Instead of using
+dedicated hardware for compressing and decompressing data from memory, this
+approach uses existing hardware that is not being utilized. Another advantage of
+not using dedicated hardware is the ability to change compression algorithms
+flexibly, since different applications have varying access patterns that are
+suited for different compression algorithms. Another use for assist warps is in
+memoization, in which frequently- computed values from main memory are cached
+and used as lookup tables, reducing the number of loads and stores required. One
+final use suggested by the authors is using an idle memory pipeline to prefetch
+values during computation to reduce the memory pressure.
+
+CABA contains a mix of hardware and software. While software-only or hardware-
+only options are possible, the former induces high overhead and complicates
+communication between primary and helper warps, and the latter makes having a
+flexible interface more difficult. The combination suggested by the authors
+dynamically inserts instructions into the execution stream in the form of assist
+warps. This requires stores of commonly used assist warps, a controller to
+trigger and manage assist warp execution, and an assist warp buffer inside the
+instruction buffer that holds triggered warps until they are scheduled. Helper
+threads share context with the parent warp, and register values in helper
+threads are not preserved between invocations, reducing the overhead of this
+technique and does not require any additional resources.
+
+The authors report a reduction in memory bandwidth of 2.1x, resulting in an
+increase in performance of 41.7\%. This is only 2.8\% less than the idea case in
+which there would be no overhead of assist warps. These helper threads also
+reduced the overall system energy by 22.2\%, which is 4.0\% over the ideal
+effect. Power consumption is also increased by 2.9\%, but this is a small price
+to pay for the performance increases.
+
+\subsection{Dynamic Thread Block Launch}
+The authors of Dynamic Thread Block Launch (DTBL) suggest a somewhat similar
+approach, this time launching thread blocks dynamically instead of assist warps
+\cite{DynamicThreadBlockLaunch}. The motivation behind this technique is that
+while emerging general-purpose applications do not have simple, regular
+structure that is easy to exploit at the highest level, they still do contain
+``pockets'' of parallelism at smaller levels. The authors explain that these
+pockets can be exploited with existing architecture using device-side nested
+kernel launching, but at a high cost. For example, CUDA supports parent kernels
+launching child kernels through device-side API calls. These child kernels can
+start at any time after being launched by their parent. This method does exploit
+pockets of parallelism, but at a high cost. On average, this CUDA method results
+in a 36.1\% decrease in performance, and also consumes significant amounts of
+memory by storing pending child kernels.
+
+The problems with existing implementations boil down to four characteristics:
+\begin{description}
+\item [High kernel density] Kernel launches are expensive, and typical
+applications currently require many kernels. For example, a typical breadth-
+first search requires about 3,000 device kernel launches, resulting in a high
+overhead and memory footprint.
+\item [Low compute intensity] While many kernels are launched using these
+methods, they are usually fine-grained and do not contain high amounts of
+parallelism. The average number of threads in these kernels is around 40, which
+is close to the warp size and could be better served by a more fine-grained
+approach.
+\item [Workload similarity] The operations performed by each dynamically
+launched kernel are usually similar, but contain different levels of parallelism
+due to differing configurations or data.
+\item [Low concurrency/scheduling efficiency] Due to the low compute intensity
+of the device kernels, multiple kernels should be able to execute in parallel.
+However, GPU architectures place limits on the number of different kernels that
+can execute concurrently, and since each kernel has a relatively small number of
+warps within it, the SMs are underutilized and struggle to hide memory
+latencies.
+\end{description}
+
+DTBL is a more fine-grained approach. Instead of launching entire kernels,
+individual thread blocks are created and added to the list of existing thread
+blocks within a kernel. These thread blocks are coalesced into an aggregated
+kernel and executed together. The existing GPU microarchitecture is extended to
+support these additional thread blocks by adding data structures to keep track
+of dynamically formed groups of thread blocks and associating them with kernels.
+Fortunately, this organization is transparent to much of the rest of the
+architecture, including the warp schedulers and control divergence mechanism.
+
+The extra data structures of this technique only contribute an area overhead of
+0.5\%. Additionally, the authors report a 1.21x performance increase over
+original GPU implementations and 1.40x over implementations using existing
+device-kernel launches.
+
+\section{Dynamic Scheduling} \label{sec:scheduling}
+\subsection{Coordinated warp scheduling (CAWA)}
+
+\subsection{Variable Warp Size (VWS)}
 This section left as an exercise to the reader
-\subsection{Coordinated warp scheduling (CAWA)}
-This section left as an exercise to the reader
-\subsection{Dynamic Thread Block Launch (DTBL)}
-This section left as an exercise to the reader
-\subsection{Equalizer}
-This section left as an exercise to the reader
+
 \subsection{Fine-grain Hardware Worklists (HWWL)}
 This section left as an exercise to the reader
-\subsection{Variable Warp Size (VWS)}
+
+\subsection{Equalizer}
 This section left as an exercise to the reader
@@
 This section left as an exercise to the reader
+
 \subsection{Managing Concurrency in Heterogeneous (CPU/GPU NoC contention)}
@@
 This section left as an exercise to the reader
+
 \subsection{Locality Aware Mapping}
 This section left as an exercise to the reader
+
 \subsection{(x) WeakBehaviorProgrammingAssumptions}

### 2015-12-08 FinalProject/Tex/paper/paper.tex
added clarifying wording in introduction
removed contractions

@@
 parallel nature of GPUs. Using hundreds or even thousands of ``cores" to perform
-computations in parallel is tempting (even if the cores aren't really cores at
+computations in parallel is tempting (even if the cores are not really cores at
 all), but brings with it a new set of challenges.
@@
 traditional desktop multiprocessing system. Even when under a heavy compute
-load, it isn't always easy to keep all of the cores doing useful work in any
+load, it is not always easy to keep all of the cores doing useful work in any
 given cycle. Between limitations in structural units, dependences between
@@
 increase performance. Section \ref{sec:scheduling} deals with more efficient
-scheduling within and among streaming multiprocessors. Section \ref{sec:memory}
-talks specifically about reducing memory contention. Section \ref{sec:software}
-is a discussion of a few software and compiler solutions to the challenges of
-GPGPU. Section
+scheduling within and among streaming multiprocessors, as well as tuning
+hardware to match the current load of the GPU. Section \ref{sec:memory} talks
+specifically about reducing memory contention. Section \ref{sec:software} is a
+discussion of a few software and compiler solutions to the challenges of GPGPU.
+Section
 \ref{sec:conclusions} concludes the survey.
@@
 ``wavefront''. Maxwell supports up to 64 warps resident per SM. Note that this
-doesn't mean that all 64 warps are executing at the same time, but that the
+does not mean that all 64 warps are executing at the same time, but that the
 architecture can support that many warp contexts concurrently. The SM maintains

### 2015-12-08 FinalProject/Tex/paper/paper.tex
finished dynamic scheduling subsection

@@
 \section{Dynamic Scheduling} \label{sec:scheduling}
-\subsection{Coordinated warp scheduling (CAWA)}
+\subsection{Critical Warps}
+GPUs manage to hide latency by swapping out stalled warps, but current
+performance is not good enough. This latency hiding is limited by the following
+characteristics of heterogeneous applications: imbalanced workloads in parallel
+warps, diverging branch behavior, irregular memory access patterns, and shared
+cache contention \cite{CoordinatedWarpScheduling}. These imbalances and
+irregularities result in differences in execution time of parallel warps. The
+slowest warp in a group is called the ``critical'' Since warps are blocked for
+synchronization purposes until all of them are finished, this needlessly extends
+the life of all the other warps and limits the performance of the application as
+a whole. The authors of Criticality-aware warp acceleration (CAWA) demonstrate
+that the average difference in execution time between the fastest and slowest
+warp is 45\% and can reach as high as 70\%.
 
-\subsection{Variable Warp Size (VWS)}
-This section left as an exercise to the reader
+CAWA seeks to reduce the disparity among execution time in parallel warps using
+three components:
+\begin{description}
+  \item[Dynamic warp criticality prediction] This predictor attempts to identify
+  the critical warp that is used in the next two components. This is
+  accomplished by using a per-warp criticality counter, which is based on the
+  degree of instruction count disparity, a measure of how much diverging
+  branches differ, as well as the number of stall cycles between two consecutive
+  instructions. These two values are combined into a single criticality counter.
+  \item[A greedy criticality-aware warp scheduler] This scheduler is built on
+  top of the two existing warp scheduling policies, and gives more time to
+  critical warps (based on the criticality counter described above). Ties are
+  broken by the age of the warp, giving priority to older warps.
+  \item[Criticality-aware cache prioritization] This scheme prioritizes cache
+  accesses. It reserves a fixed number of L1 data cache ways for critical warps.
+  The best number of ways to reserve was determined by experimental analysis and
+  found to be 8 out of 16. The cache prioritization also uses a signature-based
+  hit predictor to predict whether a given cache line belongs to the critical
+  warp.
+\end{description}
 
-\subsection{Fine-grain Hardware Worklists (HWWL)}
-This section left as an exercise to the reader
+Using an oracle to identify the critical warp could yield a 21\% performance
+improvement on average. The use of CAWA resulted in an average improvement of
+9.2\% across all applications. Applications that are more sensitive to
+scheduling policies and cache accesses saw a 23\% improvement in performance.
+The accuracy of the critical warp identifier was also measured. It successfully
+identified the critical warp as a slow warp 73\% of the time, where slow warp
+here means it is in the slower half.
 
-\subsection{Equalizer}
-This section left as an exercise to the reader
+\subsection{Variable Warp Size}
+While some of the past sections have discussed added new execution objects to
+perform work, this technique pertains to splitting and re-merging existing
+warps. There are two types of divergence in GPU execution: control flow
+divergence and memory divergence. The former deals with complex control flow in
+large warps resulting in performance loss, while the latter is related to how
+well memory requests are coalesced into single warps. When warps are too small,
+horizontal locality of memory accesses is lost; this effect is termed ``memory
+convergence slip.''
+
+In addition, workloads can be classified into three groups depending on how
+their performance is affected by warp size. Divergent warps perform poorly as
+warp size increases, due to control flow divergence as explained above. Warp-
+size insensitive warps see little to no performance changes with warp size.
+Convergent warps actually see better performance as warp size increases, due to
+coalescing execution and memory accesses into a single warp. Increasing warp
+size is better for the latter two types, since larger warps have better energy
+efficiency.
+
+Recent applications are increasingly divergent
+\cite{VariableWarpSize}: raytracing, molecular dynamics simulations, game
+physics simulations, and graph processing are all common applications for GPGPU
+and all contain many divergent warps. To combat this, warp sizes can be adjusted
+dynamically. A minimum of 4 is a good floor value, since graphics workloads
+commonly process ``quads'' (threads in groups of 4) and there are diminishing
+returns for warp sizes any smaller than this. This technique works by splitting
+the traditional datapath into 8 slices, where each slice can fetch, decode, and
+issue independently. Slices are statically assigned threads in a linear fashion
+(0-3 to slice 0, 4-7 to slice 1, etc.), and threads cannot move between slices.
+
+Allowing for variable warp sizes does not alter the register file and requires
+no changes to memory. A warp ganging unit is added that fetches and decodes once
+for all the warps in a gang. The gang issue scheduler enforces lock-step
+execution of all the slices within a gang. Independent warps can be added to
+fill holes in gangs, and gangs can be split and reformed. Fetching and issuing
+gives priority to gangs to ensure the maximum level of lockstep.
+
+The authors of this technique varied gang scheduling policies, reformation
+policies, splitting policies, and size. They found that inelastic variable warp
+sizing, in which gangs are split based on their control path and are never
+merged back together, had the best performance. They measured a 35\% increase in
+performance on a 32-wide machine, using 8-wide warps, at the cost of only 2.5\%
+area.
+
+\subsection{Hardware Worklists}
+There are two common approaches to mapping graph algorithms to GPUs. The first
+is topology-driven, in which the current thread index is used to determine on
+which nodes to operate. This approach launches a new kernel for every iteration
+and visits all nodes. Since launching kernels is expensive and some or many
+nodes do not have any useful work to do in a given iteration, this results in
+inefficiencies. The other approach is data-driven, where shared software
+worklists (SWWL) are used to determine which nodes have useful work for a given
+iteration. Only one kernel is launched at the start, and it is refilled with
+work as necessary each iteration. This approach avoids all of the useless work
+found in the topology-driven method, but can require a very large worklist,
+which makes it prone to memory contention or access irregularity. Because of
+this drawback, data-driven mapping is impractical unless aggressive software
+optimizations are made.
+
+There are four weaknesses that data-driven optimizations attempt to improve:
+memory contention on pulls, contention on pushes, load balancing, and work
+efficiency. Double-buffering is the simplest optimization and reduces memory
+contention by partitioning the worklist push and pull sections. Work chunking
+reduces push contention by grouping multiple pushes into one operation,
+amortizing the cost. Work donating works by donating excess work to other
+threads within the same block, improving load balance. Variable kernel
+configuration adjusts the number of threads spawned by a kernel call based on
+the number of active nodes in the worklist.
+
+Even with such optimizations, the data-driven approach is not ideal due to
+remaining memory contention on pushes and poor load balancing. Using fine-
+grained hardware worklists (HWWL) instead of software worklists is one method of
+dealing with such drawbacks
+\cite{UsingFineGrainHardwareWorklists}. This works by exposing the HWWL to the
+software through simple additions to the ISA. This way an application utilizing
+a SWWL can be easily adapted to use a HWWL. The microarchitecture of the GPU is
+modified to support a HWWL, distributing work to each lane in the GPU. To enable
+better load balancing, empty banks pull from other banks instead of exiting.
+There are two policies for when to redistribute work: interval-based and on-
+demand. The former periodically redistributes work at a given sampling interval.
+This has the advantage of better load balancing, but a higher cost. The
+alternative, on- demand redistribution, only redistributes when a bank pulls or
+pushes. This saves energy but has poorer load balancing than the interval-based
+policy.
+
+In addition, there are three possible ways to redistribute the work. The first
+is threshold-based, where greedy (more full) banks donate to the needy (less
+full) banks. The second scheme is local sorting-based, in which each HWWL bank
+indicates how much work it has. A work redistribution unit (WRU) in each SM
+sorts its own banks and attempts to redistribute within itself. The global
+sorting-based scheme is similar to the second, but uses a single WRU for the
+entire GPU. This last approach is unrealistic due to scalability issues.
+
+The results of this technique are promising. When compared to the SWWL, the HWWL
+had a 67\% reduction in memory stalls and between a 1.2-2.4x increase in
+performance. The area overhead of the HWWL banks was 2.5\% of the register file
+core, which is a small price to pay.
+
+\subsection{Dynamic Frequency and Parallelism Adjustment}
+The last technique in this section deals with dynamically monitoring resource
+requirements and adjusting the SM frequency, memory frequency, and number of
+threads executing. A majority of GPU applications are bottlenecked by either a
+lack of compute resources, memory bandwidth, or a limited data cache, and so
+executing the maximum number of threads and the highest frequencies is not
+always beneficial to performance. If a given resource is bottle-necked, the
+other resources become underutilized. For example, the NVIDIA Fermi architecture
+has 48KB of data cache. When spread out over the maximum number of threads that
+can execute at the same time, this leaves only 30 bytes for each.
+
+If the characteristics of an application's workload were known before execution,
+the GPU could be tuned to execute the optimal number of threads at the optimal
+SM and DRAM frequencies. Unfortunately, this is often not possible: workloads
+are heavily dependent on the given application, its input, and the specific GPU.
+Equalizer \cite{EqualizerTuningOfResources} is a runtime system that runs on
+individual SMs, determines the resource requirements of kernels, and adjusts the
+number of executing threads and frequencies of the SM and memory.
+
+Equalizer determines the overall resource requirements of a kernel based on the
+status of the warps within the kernel. Warps are classified into one of four
+groups: active warps, which are not paused; warps that cannot be executed
+because they are waiting on operands from memory; warps ready to issue to the
+memory pipeline but that the Load/Store queue cannot accept this cycle; and
+warps ready to issue to the arithmetic pipeline, but cannot because the pipeline
+is saturated. Based off these classifications, Equalizer determines whether to
+adjust any of the three parameters once per ``epoch'' (end of each execution
+window). The number of thread blocks is only changed if it is appropriate to do
+so for three consecutive epochs. New voltage and frequency values are submitted
+to the Frequency Manager every epoch and adjusted if necessary. This manager
+uses one of three discrete settings, where the low and high settings represent a
+15\% reduction or increase, respectively. This model assumes a linear change in
+voltage with frequency.
+
+Equalizer operates in one of two modes: energy efficient or high performance.
+The authors found that there were insignificant area and energy overheads from
+additional hardware needed to implement Equalizer. In performance mode, an
+overall 22\% performance boost was measured, while also saving 6\% energy. The
+efficient mode saved 15\% energy and resulted in a 5\% performance boost.
 

### 2015-12-08 FinalProject/Tex/paper/paper.tex
finished summarizing techniques

@@
 \begin{description}
-  \setlength\itemsep{1em}
+  \setlength\itemsep{0.5em}
   \item[A large register file] The register file is shared across all threads
@@
 \begin{description}
+  \setlength\itemsep{0.5em}
   \item [High kernel density] Kernel launches are expensive, and typical
@@
 \begin{description}
+  \setlength\itemsep{0.5em}
   \item[Dynamic warp criticality prediction] This predictor attempts to identify
@@
 
-\section{Reducing Memory Contention} \label{sec:memory}
-\subsection{Warped Compression}
-This section left as an exercise to the reader
+\section{Compiler and Software Techniques} \label{sec:software}
+\subsection{Divergence Management}
+We have seen that complex control flow leads to divergent branches and as a
+result divergent warps. Vector machines commonly handle this divergence
+explicitly by using predication registers to conditionally execute statements.
+Intel MIC accelerators, Cray-1 vector processors, and even AMD GPUs operate this
+way \cite{DivergenceManagment}. However, NVIDIA GPUs manage divergence
+implicitly. The GPU chooses one path and defers the execution of the other by
+pushing the current PC and a thread mask onto the ``divergence stack.''
+Reconvergence is then handled by popping off of the stack. This structure allows
+for correct nesting of divergence and reconvergence. NVIDIA also allows for
+software management of divergence using predicate registers and ``consensual
+branches.'' These branches are only taken if all the threads within the current
+warp have the same predication value. However, this functionality used very
+limitedly, only when heuristics indicate a potential performance boost.
 
-\subsection{Managing Concurrency in Heterogeneous (CPU/GPU NoC contention)}
-This section left as an exercise to the reader
+Compiler-driven approaches to divergence management exist for vector-style
+architectures, but the problem is different when applied to GPUs. Traditional
+compiler algorithms for predication are thread-agnostic, since they only
+optimize control flow for a single thread. Data-parallel architectures such as
+those in GPUs, on the other hand, must be thread-aware. One such set of
+algorithms has been developed \cite{DivergenceManagment}.
 
-\section{Compiler/SW-only design} \label{sec:software}
-\subsection{Divergence Management}
-This section left as an exercise to the reader
+These compiler algorithms remove control flow by linearizing execution paths.
+This is done by first constructing thread-aware control dependence graphs
+(CDGs). Predicate nodes are inserted between dependent basic blocks. The CDGs
+are augmented by the compiler by adding three kinds of runtime masks: loop masks
+track which threads are still executing; continue masks track which threads are
+in a current iteration; and exit masks track which threads exited by specific
+exit blocks. It is important to note that the resulting CDG is still valid for
+processing, so later compiler passes can continue to process cyclic regions.
+
+The algorithms also perform a few optimizations. If a branch condition is
+determined to be thread-invariant, the branch is replaced with a consensual
+branch and the predication is removed. This removes unnecessary work and
+lightens register pressure. Branch conditions that cannot be proven thread
+invariant have dynamic predicate uniformity tests inserted.
+
+To linearize the control flow, loops are analyzed with structural analysis.
+After the analysis has determined control flow, the direction of the edges in
+the CFG are reversed, and the CFG is processed in a depth-first post- order
+traversal. This yields a valid schedule.
+
+Applying these algorithms results in code that is 2.7\% slower on average, with
+little to no area, power, or energy implications. The main motivation for this
+approach is that is allows for less hardware design complexity and verification
+costs. From the programmer's point of view, there is no difference. In addition,
+using divergence stack control sometimes results in nondeterministic order of
+execution. Using the variance analysis in this approach allows the compiler to
+make stronger guarantees at compile time.
 
 \subsection{Locality Aware Mapping}
-This section left as an exercise to the reader
+GPUs are good at exploiting simple parallelism, and high-level language
+extensions have been developed to make writing parallel programs easier, but
+nested parallelism is still quite difficult to manage. There are multiple
+strategies for mapping nested patterns onto GPUs, but they all perform poorly
+under certain circumstances. Three such mappings are:
+\begin{description}
+  \setlength\itemsep{0.5em}
+  \item[1D mapping] Only parallelize the outer loop
+  \item[Thread-block/thread mapping]assign each iteration of the outer pattern
+  to a thread block and parallelize the inner pattern by threads within the
+  block
+  \item[Warp-based mapping] assign each iteration of the outer pattern to a warp
+  and inner iterations to threads within a warp
+\end{description}
+None of these approaches is optimal across all applications. Instead, a
+generalized mapping with strong support for nested mapping is needed. One such
+approach has been developed \cite{LocalityAwareMappingOfNestedParallelPatterns}.
 
-\subsection{(x) WeakBehaviorProgrammingAssumptions}
-This section left as an exercise to the reader
+This locality-aware mapping has its own intermediate representation (IR) that
+allows simple nesting of parallel patterns. These mappings contain three
+parameters: dimension, a dimension for each nest level; block size, the number
+of threads in a given block; and degree of parallelism (DOP), the number of
+parallel computations enabled by the mapping. Analysis of these mappings is
+flexible and not limited to one or two levels like the other mappings mentioned
+above.
+
+Two orthogonal categories are used to evaluate the strength of these mappings:
+local vs. global and hard vs. soft. Local constraints are applied to a specific
+pattern, such as having sequential accesses map to the fastest varying
+dimension. Global constraints consider multiple patterns together. Hard
+constraints must be satisfied to ensure correct execution, such as a limit on
+the number of threads within a block. Soft constraints provide performance
+hints, but can be violated if necessary. An example of a soft limit is avoiding
+thread divergence. Using these constraints, mappings are ranked, and the most
+efficient is chosen. Only candidates that meet hard constraints are considered,
+and ties are broken in favor of higher DOP.
+
+The most efficient mapping is then used to generate CUDA code. A few
+optimizations can be applied to this generation, such as preallocating memory
+before kernel launch if all loop iterations have a constant size allocation.
+Another optimization uses prefetching for multiple loop iterations when loops
+are imperfectly nested.
+
+The efficiency of these mappings varies depending on the application. For simple
+nested patterns, the generated code is 20\% slower than manually-optimized using
+the 1D mapping. However, using the fixed 2D strategies mentioned are 1.5-9.6x
+slower than the multi-dimensional mappings. On real-world applications, these
+custom mappings are 1.18x-4.38x faster than running on an 8-core CPU and
+4.5x-8.95x faster than using a 1D mapping.
+
+\section{Other Research} \label{sec:memory}
+\subsection{Warp Data Compression}
+To increase the breadth of this survey, two additional topics are discussed that
+do not fit well in the categories so far. The first is compression and
+decompression of register file accesses on a warp level. GPU register files
+contribute up to 15\% of the total GPU chip power, so there are significant
+savings available if the register file can be made more power efficient. One
+past approach has been turning off unused parts of the register file. This saves
+both leakage and dynamic power. Another approach takes advantage of register use
+locality with register file caches.
+
+The technique discussed here is compressing at the warp level
+\cite{WarpCompression}. Some GPU architectures already use data compression. For
+example, NVIDIA's Maxwell architecture claims to improve effective bandwidth by
+25\% using a compression scheme. For these applications, compression schemes
+should satisfy two conditions: the data should have a high compression ratio
+using the selected scheme, and the latency of compressing/decompressing should
+be short. In the past, most schemes have focused on the L2 or last level cache
+to avoid latency issues.
+
+To compress the values used in warps, value similarity is key. This is the
+arithmetic distance between register values. Values across threads in a warp are
+typically similar, since data arrays are commonly accessed using the thread IDs
+of successive threads, and the dynamic range of input data is small for many
+things. Power can be saved by avoiding the repeated storing, reading, and
+writing of redundant data. Distances between successive thread registers are
+calculated and placed into one of four bins: zero distance, which means the two
+values are the same; 128 bin, where the values differ by at most 128; 32K bin,
+where they differ by at most 32K, and the random bin, which holds all values
+that do not fall into any of the first three.
+
+Keeping in mind the trade-off of compression ratio vs latency, the base-delta-
+immediate (BDI) scheme is most suitable warp compression. This scheme uses a
+base value, and subsequent deltas are added to the base to determine each
+subsequent value. Computed values are compressed before writing to the register
+file and decompressed before reading. Since most warp instructions have two
+sources and one destination, it makes sense to have two decompression units and
+a single compression unit to process one warp per cycle. Since modern GPU
+architectures can process two warps per cycle, the number of each of these is
+doubled, for a total of four decompression units and two compression units.
+
+The results of this compression scheme are promising. The area overhead is only
+0.3\% of the total SM size, and an overall 25\% power reduction (35\% dynamic,
+10\% static) is achieved in the register file. While added pipeline stages for
+compression and decompression add time, the total performance loss is only
+0.1\%. Warp compression is clearly a promising technique for reducing power in
+the massive register files of GPUs.
+
+\subsection{Memory Contention in Heterogeneous Multiprocessing Systems}
+Modern computing system architectures include both a CPU and GPU in a master-
+slave configuration, in which the CPU offloads heavy parallel computations to
+the GPU. In the future, there will likely be many CPUs and GPUs operating in
+conjunction within a single system. These network-on-chip (NoC) configurations
+are called heterogeneous multiprocessing architectures
+\cite{ManagingConcurrencyInHeterogeneous}.
+
+CPU applications tend to be more latency sensitive, while GPU applications are
+more bandwidth sensitive. In such heterogeneous architectures containing both,
+GPUs tend to cannibalize the performance of CPUs much more than the other way
+around. The goal of the technique described here is to prevent this
+cannibalization and improve the performance of both kinds of computation. This
+technique assumes a tile-based mesh architecture consisting of both CPUs and
+GPUs. The CPU to GPU ratio is 2:1 because a single GPU SM occupies roughly half
+the space of a CPU core. The last level cache of the system is shared by CPUs
+and GPUs, and they use a shared network and shared memory channels.
+
+GPU concurrency can have a positive or negative effect on other GPU performance.
+Increasing the number of concurrent warps can help performance but can also
+thrash the cache or cause memory congestion. The effects of GPU concurrency on
+CPU performance, however, are much more negative. Memory congestion hurts CPUs
+much more than GPUs, as they do not have enough thread-level parallelism to hide
+the latency.
+
+Two schemes are presented to keep GPUs from monopolizing the resources of the
+shared system. The first is CPU-Centric Concurrency Management (CM-CPU). This
+scheme dynamically monitors congestion and modulates the number of active warps
+on GPUs in steps of two. The second scheme is Balanced Concurrency Management
+(CM-BAL), which detects if GPUs do not have enough latency tolerance, and
+increases the number of active warps to hide it. The goal of this scheme is to
+recover performance lost from the first scheme.
+
+These approaches only require 16 bytes of overhead for the GPU core, 3 bytes per
+memory partition, and 8 bytes for the thread-block scheduler. CM-CPU resulted in
+an 11\% average GPU performance loss and a 24\% CPU performance increase. CM-BAL
+resulted in a 7\% loss in GPU performance and a 7\% gain in CPU performance.
+Overall speedup of the system depends on the type of application and relative
+importance of CPU and GPU computation. CM-CPU ranges from 0.875x-1.25x overall
+speedup, which CM-BAL yields a more balanced 1.05x-1.1x.
 

### 2015-12-08 FinalProject/Tex/paper/paper.tex
Added conclusion

@@
 \section{Conclusions} \label{sec:conclusions}
-\begin{itemize}
-\item Mention what the survey talked about
-  \begin{itemize}
-  \item Background
-  \item Overview
-  \item Challenges
-  \item Designs
-  \end{itemize}
-\item Explain changes occurring in field
-\item Talk about what the future could hold
-\end{itemize}
+This survey article gave a brief overview of GPU architecture, general-purpose
+computing on GPUs and the corresponding challenges, and current areas of
+research in GPGPU.
+
+GPGPU is a relatively young field but grows more popular each year. The main
+struggles with GPGPU are maintaining high utilization of resources in the face
+of complicated applications that are not easily mapped to GPU architectures
+while keeping contention of other resources, especially those related to memory.
+High memory pressure prevents GPUs from hiding memory latency and seriously
+degrades performance.
+
+Techniques within GPGPU continue to be developed focusing on a number of
+different areas. While this survey has only scratched the surface, it has
+provided a glimpse into current efforts to dynamically create helper warps and
+thread-blocks, dynamically schedule warps and adjust the state of the SMs on the
+fly, and use better compiler techniques to avoid divergence and produce better
+mappings to GPU architectures. A couple other novel research areas such as data
+compression and reducing memory contention were also discussed.
+
+GPUs may have started out as specialized hardware for rendering graphics, but
+they have begun to be used on much more diverse problems. If the challenges of
+GPGPU can be overcome, significant performance improvements across a wide range
+of applications could be unlocked.
 

### 2015-12-08 FinalProject/Tex/paper/paper.tex
fixed various grammatical and line break issues

@@
 applying this architecture to heterogeneous problems are discussed. Several
-different current areas of research are explored, dynamic warp scheduling and
-hardware tuning, reducing memory contention, and software solutions to the
-challenges of GPGPU. While the challenges presented are difficult to overcome,
-general-purpose computing on GPUs continues to make progress in the areas of
-performance, energy, and power.
+different current areas of research are explored, including dynamic warp
+scheduling and hardware tuning, reducing memory contention, and software
+solutions to the challenges of GPGPU. While the challenges presented are
+difficult to overcome, general-purpose computing on GPUs continues to make
+progress in the areas of performance, energy, and power.
 \end{abstract}
@@
 traditional desktop multiprocessing system. Even when under a heavy compute
-load, it is not always easy to keep all of the cores doing useful work in any
-given cycle. Between limitations in structural units, dependences between
+load, it is not easy to keep all of the cores doing useful work in any given
+cycle. Between limitations in structural units, dependences between
 instructions, and latencies of memory accesses, the processor cannot help but
@@
 \subsection{Streaming Multiprocessors}
-Modern unified GPUs consist of multiple streaming multiprocessors (SMs). While
-there are many GPU architectures available today, the following description is
-based off of NVIDIA's Maxwell architecture, which was released in 2014. These
-SMs are similar to SIMD processors and typically contain the following
-components:
+Modern unified GPUs consist of multiple streaming multiprocessors (SMs). Figure
+\ref{fig:baseline} shows a baseline GPU configuration
+\cite{DynamicThreadBlockLaunch}. While there are many GPU architectures
+available today, the following description is based off of NVIDIA's Maxwell
+architecture, which was released in 2014. These SMs are similar to SIMD
+processors and typically contain the following components:
 \begin{description}
@@
 Grids are what actually run on the GPU, and they are composed of multiple
-``thread- blocks'', also called a cooperative thread array (CTA) outside of
-NVIDIA's nomenclature. Maxwell supports up to 64K blocks within a grid. Thread-
-blocks contain potentially many threads, on the order of 1024 for modern
+``thread-blocks'', also called a cooperative thread array (CTA) outside of
+NVIDIA's nomenclature. Maxwell supports up to 64K blocks within a grid.
+Thread-blocks contain potentially many threads, on the order of 1024 for modern
 architectures. Each thread within a thread-block has the same life-cycle as all
@@
 such that all threads execute the same instruction on different data, just a
-like a SIMD unit. With this in mind, it would be more accurate to simply call
+like a SIMD unit. With this in mind, it would be more accurate to call
 ``threads'' a sequence of SIMD lane instructions, but this survey will continue
 to refer to them as threads to maintain consistency with the literature. The
-term ``Warp'' is the term used by NVIDIA, while AMD and OpenCL use
+term ``warp'' is the term used by NVIDIA, while AMD and OpenCL use
 ``wavefront''. Maxwell supports up to 64 warps resident per SM. Note that this
@@
 the warp schedulers from this pool.
+
+\begin{figure}
+  \centering
+  \includegraphics[]{BaselineGPU.png} \caption{Test caption} \label{fig:baseline}
+\end{figure}
 
@@
 As we have seen, GPUs are capable of supporting thousands of threads, but
-bottlenecks in heterogeneous applications limit performance. \cite{AssistWarps}
-present the idea of using ``assist warps'' in their Core-Assisted Bottleneck
-Acceleration (CABA) scheme. These special warps are generated by dedicated
-hardware and are used as helper threads to increase performance and efficiency.
-Assist warps share the same context as the parent warp and uses some of the
-available registers, since investigation shows that 24\% of the register file is
-unallocated on average.
+bottlenecks in heterogeneous applications limit performance. Core-Assisted
+Bottleneck Acceleration \cite{AssistWarps} is a scheme that uses ``assist
+warps.'' These special warps are generated by dedicated hardware and are used as
+helper threads to increase performance and efficiency. Assist warps share the
+same context as the parent warp and uses some of the available registers, since
+investigation shows that 24\% of the register file is unallocated on average.
 
@@
 flexibly, since different applications have varying access patterns that are
-suited for different compression algorithms. Another use for assist warps is in
-memoization, in which frequently- computed values from main memory are cached
-and used as lookup tables, reducing the number of loads and stores required. One
+suited for different compression algorithms. Assist warps can also be used for
+memoization, in which frequently-computed values from main memory are cached and
+used as lookup tables, reducing the number of loads and stores required. One
 final use suggested by the authors is using an idle memory pipeline to prefetch
@@
 
-CABA contains a mix of hardware and software. While software-only or hardware-
-only options are possible, the former induces high overhead and complicates
-communication between primary and helper warps, and the latter makes having a
-flexible interface more difficult. The combination suggested by the authors
-dynamically inserts instructions into the execution stream in the form of assist
-warps. This requires stores of commonly used assist warps, a controller to
-trigger and manage assist warp execution, and an assist warp buffer inside the
-instruction buffer that holds triggered warps until they are scheduled. Helper
-threads share context with the parent warp, and register values in helper
-threads are not preserved between invocations, reducing the overhead of this
-technique and does not require any additional resources.
+CABA contains a mix of hardware and software. While software-only or
+hardware-only options are possible, the former induces high overhead and
+complicates communication between primary and helper warps, and the latter makes
+having a flexible interface more difficult. The combination suggested by the
+authors dynamically inserts instructions into the execution stream in the form
+of assist warps. This requires stores of commonly used assist warps, a
+controller to trigger and manage assist warp execution, and an assist warp
+buffer inside the instruction buffer that holds triggered warps until they are
+scheduled. Helper threads share context with the parent warp, and register
+values in helper threads are not preserved between invocations, reducing the
+overhead of this technique and does not require any additional resources.
 
@@
 \cite{DynamicThreadBlockLaunch}. The motivation behind this technique is that
-while emerging general-purpose applications do not have simple, regular
+while emerging general-purpose applications do not have a simple, regular
 structure that is easy to exploit at the highest level, they still do contain
@@
 start at any time after being launched by their parent. This method does exploit
-pockets of parallelism, but at a high cost. On average, this CUDA method results
-in a 36.1\% decrease in performance, and also consumes significant amounts of
-memory by storing pending child kernels.
+pockets of parallelism, but this CUDA method results in an average of 36.1\%
+decrease in performance, and also consumes significant amounts of memory by
+storing pending child kernels.
 
-The problems with existing implementations boil down to four characteristics:
+The problems with existing implementations reduce down to four characteristics:
 \begin{description}
@@
   \item [High kernel density] Kernel launches are expensive, and typical
-  applications currently require many kernels. For example, a typical breadth-
-  first search requires about 3,000 device kernel launches, resulting in a high
-  overhead and memory footprint.
+  applications currently require many kernels. For example, a typical
+  breadth-first search requires about 3,000 device kernel launches, resulting in
+  a high overhead and memory footprint.
   \item [Low compute intensity] While many kernels are launched using these
@@
 irregularities result in differences in execution time of parallel warps. The
-slowest warp in a group is called the ``critical'' Since warps are blocked for
-synchronization purposes until all of them are finished, this needlessly extends
-the life of all the other warps and limits the performance of the application as
-a whole. The authors of Criticality-aware warp acceleration (CAWA) demonstrate
-that the average difference in execution time between the fastest and slowest
-warp is 45\% and can reach as high as 70\%.
+slowest warp in a group is called the ``critical'' warp. Since warps are blocked
+for synchronization purposes until all of them are finished, this needlessly
+extends the life of all the other warps and limits the performance of the
+application as a whole. The authors of Criticality-aware warp acceleration
+(CAWA) demonstrate that the average difference in execution time between the
+fastest and slowest warp is 45\% and can reach as high as 70\%.
 
@@
 While some of the past sections have discussed added new execution objects to
-perform work, this technique pertains to splitting and re-merging existing
-warps. There are two types of divergence in GPU execution: control flow
-divergence and memory divergence. The former deals with complex control flow in
-large warps resulting in performance loss, while the latter is related to how
-well memory requests are coalesced into single warps. When warps are too small,
-horizontal locality of memory accesses is lost; this effect is termed ``memory
-convergence slip.''
+perform work, this technique pertains to splitting and re-merging existing warps
+\cite{VariableWarpSize}. There are two types of divergence in GPU execution:
+control flow divergence and memory divergence. The former deals with complex
+control flow in large warps resulting in performance loss, while the latter is
+related to how well memory requests are coalesced into single warps. When warps
+are too small, horizontal locality of memory accesses is lost; this effect is
+termed ``memory convergence slip.''
 
@@
 their performance is affected by warp size. Divergent warps perform poorly as
-warp size increases, due to control flow divergence as explained above. Warp-
-size insensitive warps see little to no performance changes with warp size.
+warp size increases, due to control flow divergence as explained above.
+Warp-size insensitive warps see little to no performance changes with warp size.
 Convergent warps actually see better performance as warp size increases, due to
@@
 
-Recent applications are increasingly divergent
-\cite{VariableWarpSize}: raytracing, molecular dynamics simulations, game
-physics simulations, and graph processing are all common applications for GPGPU
-and all contain many divergent warps. To combat this, warp sizes can be adjusted
-dynamically. A minimum of 4 is a good floor value, since graphics workloads
-commonly process ``quads'' (threads in groups of 4) and there are diminishing
-returns for warp sizes any smaller than this. This technique works by splitting
-the traditional datapath into 8 slices, where each slice can fetch, decode, and
-issue independently. Slices are statically assigned threads in a linear fashion
-(0-3 to slice 0, 4-7 to slice 1, etc.), and threads cannot move between slices.
+Recent applications are increasingly divergent: raytracing, molecular dynamics
+simulations, game physics simulations, and graph processing are all common
+applications for GPGPU and all contain many divergent warps. To combat this,
+warp sizes can be adjusted dynamically. A minimum of 4 is a good floor value,
+since graphics workloads commonly process ``quads'' (threads in groups of 4) and
+there are diminishing returns for warp sizes any smaller than this. This
+technique works by splitting the traditional datapath into 8 slices, where each
+slice can fetch, decode, and issue independently. Slices are statically assigned
+threads in a linear fashion (0-3 to slice 0, 4-7 to slice 1, etc.), and threads
+cannot move between slices.
 
@@
 \subsection{Hardware Worklists}
-There are two common approaches to mapping graph algorithms to GPUs. The first
-is topology-driven, in which the current thread index is used to determine on
-which nodes to operate. This approach launches a new kernel for every iteration
-and visits all nodes. Since launching kernels is expensive and some or many
-nodes do not have any useful work to do in a given iteration, this results in
-inefficiencies. The other approach is data-driven, where shared software
-worklists (SWWL) are used to determine which nodes have useful work for a given
-iteration. Only one kernel is launched at the start, and it is refilled with
-work as necessary each iteration. This approach avoids all of the useless work
-found in the topology-driven method, but can require a very large worklist,
-which makes it prone to memory contention or access irregularity. Because of
-this drawback, data-driven mapping is impractical unless aggressive software
-optimizations are made.
+There are two common approaches to mapping graph algorithms to GPUs
+\cite{UsingFineGrainHardwareWorklists}. The first is topology-driven, in which
+the current thread index is used to determine on which nodes to operate. This
+approach launches a new kernel for every iteration and visits all nodes. Since
+launching kernels is expensive and some or many nodes do not have any useful
+work to do in a given iteration, this results in inefficiencies. The other
+approach is data-driven, where shared software worklists (SWWL) are used to
+determine which nodes have useful work for a given iteration. Only one kernel is
+launched at the start, and it is refilled with work as necessary each iteration.
+This approach avoids all of the useless work found in the topology-driven
+method, but can require a very large worklist, which makes it prone to memory
+contention or access irregularity. Because of this drawback, data-driven mapping
+is impractical unless aggressive software optimizations are made.
 
@@
 Even with such optimizations, the data-driven approach is not ideal due to
-remaining memory contention on pushes and poor load balancing. Using fine-
-grained hardware worklists (HWWL) instead of software worklists is one method of
-dealing with such drawbacks
-\cite{UsingFineGrainHardwareWorklists}. This works by exposing the HWWL to the
+remaining memory contention on pushes and poor load balancing. Using
+fine-grained hardware worklists (HWWL) instead of software worklists is one
+method of dealing with such drawbacks. This works by exposing the HWWL to the
 software through simple additions to the ISA. This way an application utilizing
@@
 better load balancing, empty banks pull from other banks instead of exiting.
-There are two policies for when to redistribute work: interval-based and on-
-demand. The former periodically redistributes work at a given sampling interval.
-This has the advantage of better load balancing, but a higher cost. The
-alternative, on- demand redistribution, only redistributes when a bank pulls or
+There are two policies for when to redistribute work: interval-based and
+on-demand. The former periodically redistributes work at a given sampling
+interval.This has the advantage of better load balancing, but a higher cost. The
+alternative, on-demand redistribution, only redistributes when a bank pulls or
 pushes. This saves energy but has poorer load balancing than the interval-based
@@
 executing the maximum number of threads and the highest frequencies is not
-always beneficial to performance. If a given resource is bottle-necked, the
-other resources become underutilized. For example, the NVIDIA Fermi architecture
-has 48KB of data cache. When spread out over the maximum number of threads that
-can execute at the same time, this leaves only 30 bytes for each.
+always beneficial to performance \cite{EqualizerTuningOfResources}. If a given
+resource is bottle-necked, the other resources become underutilized. For
+example, the NVIDIA Fermi architecture has 48KB of data cache. When spread out
+over the maximum number of threads that can execute at the same time, this
+leaves only 30 bytes for each.
 
@@
 are heavily dependent on the given application, its input, and the specific GPU.
-Equalizer \cite{EqualizerTuningOfResources} is a runtime system that runs on
-individual SMs, determines the resource requirements of kernels, and adjusts the
-number of executing threads and frequencies of the SM and memory.
+Equalizer is a runtime system that runs on individual SMs, determines the
+resource requirements of kernels, and adjusts the number of executing threads
+and frequencies of the SM and memory.
 
@@
 After the analysis has determined control flow, the direction of the edges in
-the CFG are reversed, and the CFG is processed in a depth-first post- order
+the CFG are reversed, and the CFG is processed in a depth-first post-order
 traversal. This yields a valid schedule.
@@
 Applying these algorithms results in code that is 2.7\% slower on average, with
-little to no area, power, or energy implications. The main motivation for this
+little to no effect on area, power, or energy. The main motivation for this
 approach is that is allows for less hardware design complexity and verification
@@
 savings available if the register file can be made more power efficient. One
-past approach has been turning off unused parts of the register file. This saves
-both leakage and dynamic power. Another approach takes advantage of register use
-locality with register file caches.
+past approach has been turning off unused parts of the register file, which
+saves both leakage and dynamic power. Another approach takes advantage of
+register use locality with register file caches.
 
@@
 
-To compress the values used in warps, value similarity is key. This is the
-arithmetic distance between register values. Values across threads in a warp are
-typically similar, since data arrays are commonly accessed using the thread IDs
-of successive threads, and the dynamic range of input data is small for many
-things. Power can be saved by avoiding the repeated storing, reading, and
-writing of redundant data. Distances between successive thread registers are
-calculated and placed into one of four bins: zero distance, which means the two
-values are the same; 128 bin, where the values differ by at most 128; 32K bin,
-where they differ by at most 32K, and the random bin, which holds all values
-that do not fall into any of the first three.
+To compress the values used in warps, the arithmetic distance between register
+values should be small. This is known as value similarity. Values across threads
+in a warp are typically similar, since data arrays are commonly accessed using
+the thread IDs of successive threads, and the dynamic range of input data is
+often small as well. Power can be saved by avoiding the repeated reading and
+writing of mostly redundant data. Distances between successive thread registers
+are thus calculated and used to place the registers into one of four bins: zero
+distance, which means the base register and current register are the same; 128
+bin, where the values differ by at most 128; 32K bin, where they differ by at
+most 32K, and the random bin, which holds all values that do not fall into any
+of the first three.
 
-Keeping in mind the trade-off of compression ratio vs latency, the base-delta-
-immediate (BDI) scheme is most suitable warp compression. This scheme uses a
-base value, and subsequent deltas are added to the base to determine each
-subsequent value. Computed values are compressed before writing to the register
-file and decompressed before reading. Since most warp instructions have two
-sources and one destination, it makes sense to have two decompression units and
-a single compression unit to process one warp per cycle. Since modern GPU
-architectures can process two warps per cycle, the number of each of these is
-doubled, for a total of four decompression units and two compression units.
+Keeping in mind the trade-off of compression ratio vs latency, the
+base-delta-immediate (BDI) scheme is most suitable for warp compression. This
+scheme uses one register a base value, and subsequent deltas are added to the
+base to determine each subsequent register's value. Computed values are
+compressed before writing to the register file and decompressed before reading.
+Since most warp instructions have two sources and one destination, it makes
+sense to have two decompression units and a single compression unit to process
+one warp per cycle. Since modern GPU architectures can process two warps per
+cycle, the number of each of these is doubled, for a total of four decompression
+units and two compression units.
 
@@
 \subsection{Memory Contention in Heterogeneous Multiprocessing Systems}
-Modern computing system architectures include both a CPU and GPU in a master-
-slave configuration, in which the CPU offloads heavy parallel computations to
-the GPU. In the future, there will likely be many CPUs and GPUs operating in
-conjunction within a single system. These network-on-chip (NoC) configurations
-are called heterogeneous multiprocessing architectures
+Modern computing system architectures include both a CPU and GPU in a
+master-slave configuration, in which the CPU offloads heavy parallel
+computations to the GPU. In the future, there will likely be many CPUs and GPUs
+operating in conjunction within a single system. These network-on-chip (NoC)
+configurations are called heterogeneous multiprocessing architectures
 \cite{ManagingConcurrencyInHeterogeneous}.
@@
 CPU applications tend to be more latency sensitive, while GPU applications are
-more bandwidth sensitive. In such heterogeneous architectures containing both,
-GPUs tend to cannibalize the performance of CPUs much more than the other way
-around. The goal of the technique described here is to prevent this
-cannibalization and improve the performance of both kinds of computation. This
-technique assumes a tile-based mesh architecture consisting of both CPUs and
-GPUs. The CPU to GPU ratio is 2:1 because a single GPU SM occupies roughly half
-the space of a CPU core. The last level cache of the system is shared by CPUs
-and GPUs, and they use a shared network and shared memory channels.
+more bandwidth sensitive. In such heterogeneous architectures containing both
+kinds of processors, GPUs tend to cannibalize the performance of CPUs much more
+than the other way around. The goal of the technique described here is to
+prevent this cannibalization and improve the performance of both kinds of
+computation. This technique assumes a tile-based mesh architecture consisting of
+both CPUs and GPUs. The CPU to GPU ratio is assumed to be 2:1 because a single
+GPU SM occupies roughly half the space of a CPU core. The last level cache of
+the system is shared by CPUs and GPUs, and they use a shared network and shared
+memory channels.
 
-GPU concurrency can have a positive or negative effect on other GPU performance.
-Increasing the number of concurrent warps can help performance but can also
-thrash the cache or cause memory congestion. The effects of GPU concurrency on
-CPU performance, however, are much more negative. Memory congestion hurts CPUs
-much more than GPUs, as they do not have enough thread-level parallelism to hide
-the latency.
+GPU concurrency can have a positive or negative effect on other GPU performance,
+since increasing the number of concurrent warps can help performance but can
+also cause cache thrashing and/or memory congestion. The effects of GPU
+concurrency on CPU performance, however, are much more negative. Memory
+congestion hurts CPUs much more than GPUs, as they do not have enough
+thread-level parallelism to hide the latency.
 
@@
 The author would like to thank Dr. Penry for teaching this course and for
-providing extensive design and learning opportunities in this course. It has
-been a lot of work, but this course has helped the author know that he has
-chosen the right field.
+providing extensive design and learning opportunities throughout the semester.
+It has been a lot of work, but this course has given the author a glimpse into
+what computer architecture is all about and helped him know that he has chosen
+the right field.
 \end{acks}
